### 2020-11-10 paper/paper.tex
Update

\documentclass[a4paper,12pt]{article}
\usepackage{standalone}

\usepackage[utf8]{inputenc}
\usepackage[english]{babel}
\usepackage{mathtools}
\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{amsthm}
\usepackage{mathrsfs}
\usepackage{eufrak}
\usepackage{cite}

\makeatletter

\newcommand{\redc}{\underline{\mathrm{REDC}}}

\newcommand*{\mydiv}{%
  \nonscript\mskip-\medmuskip\mkern5mu%
  \mathbin{\operator@font div}\penalty900\mkern5mu%
  \nonscript\mskip-\medmuskip
}
\newcommand*{\mymod}{%
  \nonscript\mskip-\medmuskip\mkern5mu%
  \mathbin{\operator@font mod}\penalty900\mkern5mu%
  \nonscript\mskip-\medmuskip
}
\newsavebox{\@brx}
\newcommand{\myllangle}[1][]{\savebox{\@brx}{\(\m@th{#1\langle}\)}%
  \mathopen{\copy\@brx\kern-0.5\wd\@brx\usebox{\@brx}}}
\newcommand{\myrrangle}[1][]{\savebox{\@brx}{\(\m@th{#1\rangle}\)}%
  \mathclose{\copy\@brx\kern-0.5\wd\@brx\usebox{\@brx}}}
\makeatother

\title{Speeding up decimal multiplication}
\author{Viktor Krapivensky}
\date{%
    \today
}

\begin{document}

\maketitle

\begin{abstract}
    Decimal multiplication is the task of multiplying two numbers in base $10^N.$
    Specifically, we focus on the number-theoretic transform (NTT) family of algorithms.
    Using only portable techniques, we achieve a 3x---5x speedup over the \textbf{mpdecimal} library.
    In this paper we describe our implementation and discuss further possible optimizations.
    Our contributions are the following:
    (1) the ``linearity trick'' for using Montgomery reduction with NTT without separate steps of conversion into/out of Montgomery form;
    (2) a simple cache-oblivious algorithm for in-place $2n \times n$ or $n \times 2n$ matrix transposition, the need for which arises in the ``six-step algorithm'' variation of the matrix Fourier algorithm, and which does not seem to be widely known.
    Notable findings include:
    (1) use of two prime moduli instead of three makes sense even considering the worst case of increasing the size of the input, and makes for simpler answer recovery;
    (2) the ``six-step algorithm'', supposedly fast for large transforms due to cache-friendliness, is of little use on modern systems for transforms of sizes up to $\approx 2^{27}.$
\end{abstract}

\section{Introduction}

Fast multiplication of large decimal numbers is of interest in computer algebra systems, arbitrary precision calculators like \textbf{bc}, and generally software that needs to present the result of some calculations in decimal form.
Since conversion from binary to decimal and vice versa takes $\Theta(\mathscr{M}(n) \log n)$ time, where $\mathscr{M}(n)$ is the time needed to multiply two numbers of size $n,$
it makes sense to keep the base decimal if the calculations themselves take $o(\mathscr{M}(n) \log n)$ time. Otherwise, it might be better to pay the cost of conversions and perform the calculations in binary,
as the binary form, being native for computers, has lower hidden constant. Also, as far as NTT is concerned, the binary form allows for better granularity when deciding how many ``digits'' to pack into a single word.

It is a common theme in practical analysis of algorithms that ``fancy'' algorithms are slow when $n$ is small.
For example, virtually all partical implementations of the ``fancy'' sorting algorithms that are divide-and-conquer in nature --- quick sort and merge sort --- do in fact fall back to ``dumb'' quadratic sorts if the size of the input if less than some threshold.
In some cases, it is not even a matter of ``fancy'' versus ``dumb'' dichotomy, but rather there is a ``hierarchy of fanciness'':
a group of algorithms with different asymptotic behavior such that it makes sense to use each one of them only for some range of values of $n.$
To give an example, \textbf{libgmp} employs the following ``hierarchy of fanciness'' for multiplication:
first goes the ``basecase'' (quadratic) algorithm; then, the Karatsuba algorithm; then, variations of the Toom-Cook algorithm (Toom-3; then Toom-4; then Toom-6.5; then Toom-8.5); finally, the Schönhage--Strassen algorithm (which is Fourier transform-based).

Indeed, for smaller sizes of the multiplicands, it does makes sense to use quadratic, Karatsuba or Toom-Cook algorithms;
as $n$ gets large, however, asymptotic considerations start to outweigh.
For multiplication, this means that different variations of the fast Fourier transform start to be used.

\section{Overview of the discrete Fourier transform}

The plan of this section is as follows:
\begin{itemize}

    \item First, we define the notion of cyclic convolution for an arbitrary algebraic ring.

    \item Next, we show how to reduce multiplication of two integers $x, \, y$ in base $B,$ such that
          $x \in [0; B^n - 1], \, y \in [0; B^m - 1],$
          to cyclic convolution of size $n+m$ (or of any larger size) over ring $\mathbb{Z}$ with some
          special properties: inputs to such a convolutions are in
          $[0; B-1]$
          and outputs are in
          $[0; (B-1)^2 \min\{n, m\}].$

    \item Then, we introduce discrete Fourier transform (DFT) and inverse discrete Fourier transform (IDFT)
          in their general form --- over an algebraic field $\mathcal{F}.$
          We show the connection between cyclic convolution and DFT/IDFT.
          We will learn that, employing fast Fourier transform (FFT) for DFT and IDFT, it is possible to
          compute cyclic convolutions over $\mathcal{F}$ efficiently.

    \item Finally, we discuss how different Fourier transform-based algorithms for integer
          multiplication employ different strategies to reduce the computation of cyclic
          convolutions over $\mathbb{Z}$ to cyclic convolutions over different algebraic fields.
\end{itemize}

The following notation is employed throughout this section:
$$\langle x_0, \ldots, x_{n-1} \rangle [i] = x_i.$$

By $\mathbb{N}$ we mean $\{0, 1, 2, \ldots\}.$

\subsection{Cyclic convolution}

We now define the cyclic convolution.
Consider an aribtrary algebraic ring $\mathcal{R}.$
The cyclic convolution $x \star y, \, \star \colon \mathcal{R}^n \times \mathcal{R}^n \to \mathcal{R}^n$ is defined as follows:
$$(x \star y)[k] = \sum\limits_{i=0}^{n - 1} x[i] y[(k - i) \mymod n].$$

\subsection{Multiplication via cyclic convolution over $\mathbb{Z}$}

Let us now fix a base $B \in \mathbb{N}, \, B > 1,$ and introduce the following notation:
$$\myllangle x_0, \ldots, x_{n-1} \myrrangle = \sum\limits_{i=0}^{n-1} x_i B^i,$$
where $x_i \in \mathbb{N}, \, x_i < B.$ In order words, square brackets mean zero-based tuple indexing.

Let us say we want to multiply
$x = \myllangle x_0, \ldots, x_{n-1} \myrrangle$
and
$y = \myllangle y_0, \ldots, y_{m-1} \myrrangle.$

This is to say, we want to find the product
$z = xy = \myllangle z_0, \ldots, z_{n+m-1} \myrrangle.$

We compute the following cyclic convolution (over the ring of $\mathbb{Z}$):
$$\langle c_0, \ldots, c_{n+m-1} \rangle =
\langle x_0, \ldots, x_{n-1}, \underbrace{0, \ldots, 0}_{\text{$m$ zeros}} \rangle
\star
\langle y_0, \ldots, y_{m-1}, \underbrace{0, \ldots, 0}_{\text{$n$ zeros}} \rangle.$$

Note that $c_i \le (B-1)^2 \min\{n, m\}$ as $x_i < B, \, y_i < B,$
and, because of our zero padding, the summation in the definition of the cyclic convolution
can have at most $\min\{n,m\}$ non-zero summands.

The representation of $z$ in base $B$ can then be recovered by the following iterative process:
\begin{eqnarray*}
\sigma_0 &=& 0; \\
z_i &=& (\sigma_i + c_i) \mymod B; \\
\sigma_{i+1} &=& (\sigma_i + c_i) \mydiv B.
\end{eqnarray*}

It would be useful to obtain some upper bound on the value of $\sigma_i.$

Define $L = (B-1)^2 \min\{n,m\}$ and remember $c_i \le L.$
Define also the following sequence:
$$a_0 = 0, \, a_{i+1} = (a_i + L) \mydiv B.$$

Clearly $\sigma_i \le a_i$ irrespective of the concrete values of $\langle c_0, \ldots, c_{n+m-1} \rangle.$

Define another sequence by
$$a^\prime_0 = 0, \, a^\prime_{i+1} = (a^\prime_i + L) / B.$$

Now $a^\prime_i \ge a_i$ and
$$a^\prime_i =
\sum\limits_{j=0}^{i-1} \frac{L}{B^{i - j}} =
\sum\limits_{j=1}^{i} \frac{L}{B^j} =
L \sum\limits_{j=1}^{i} \frac{1}{B^j}.$$

We have
$$\sigma_i \le a_i \le a^\prime_i < \lim\limits_{i \to \infty} a'_i = \frac{L}{B - 1} = (B-1)\min\{n, m\}.$$

We have thus reduced multiplication of $n$-digit $x$ and $m$-digit $y$ to a convolution of size
$n+m.$ But note that, for any $k \in \mathbb{N},$
$$\myllangle x_0, \ldots, x_{n-1} \myrrangle = \myllangle x_0, \ldots, x_{n-1}, \underbrace{0, \ldots, 0}_{\text{$k$ zeros}} \myrrangle,$$
so we can as well reduce it to a convolution of any size $n+k+m \ge n+m$ by padding $x$ (or, similarly, $y$)
with $k$ higher zeros and ignoring the higher $k$ elements of the resulting vector.

\subsection{Discrete Fourier transform}

We fix an algebraic field $\mathcal{F}$ and the length of transform, $N \in \mathbb{N}.$

We will pretend natural numbers are in $\mathcal{F}$ by adopting the following identity:
$$n = \underbrace{\overline{1} + \overline{1} + \cdots + \overline{1}}_\text{$n$ times},$$
where $\overline{1} \in \mathcal{F}$ is the multiplicative identity in $\mathcal{F};$ the degenerate case of this identity is $0 = \overline{0},$
where $\overline{0} \in \mathcal{F}$ is the additive identity in $\mathcal{F}.$

We require that there exists a principal root of unity of order $N$ in $\mathcal{F}$:
such $\xi \in \mathcal{F}$ that $\xi^N = 1$ and
$$\sum\limits_{i=0}^{N-1} \xi^{ik} = 0$$
for every $0 < k < N.$ We fix one such $\xi.$

We also require that $N$ be invertible (i.e., non-zero) in $\mathcal{F}.$

Discrete Fourier transform $\mathfrak{f} \colon \mathcal{F}^N \to \mathcal{F}^N$ is then defined as follows:
$$\mathfrak{f}(x)[k] = \sum\limits_{i=0}^{N-1} x[i] \xi^{ik}.$$

Inverse discrete Fourier transform $\mathfrak{f}^{-1} \colon \mathcal{F}^N \to \mathcal{F}^N$ is defined similarly,
up to the negated exponent of $\xi$ and multiplication by $N^{-1}$:
$$\mathfrak{f}^{-1}(x)[k] = N^{-1} \sum\limits_{i=0}^{N-1} x[i] \xi^{-ik}.$$

It is easy to see that both $\mathfrak{f}$ and $\mathfrak{f}^{-1}$ are linear, which means that for any
$\alpha \in \mathcal{F}$ and $x \in \mathcal{F}^N,$
$\mathfrak{f}(\alpha x) = \alpha \mathfrak{f}(x)$
and
$\mathfrak{f}^{-1}(\alpha x) = \alpha \mathfrak{f}^{-1}(x).$ Here, the product of a scalar and a vector
$\alpha v = \alpha \langle v_0, \ldots, v_{n-1} \rangle$ denotes, as usual,
$\langle \alpha v_0, \ldots, \alpha v_{n-1} \rangle.$

The convolution theorem says that, for any two vectors $x, y \in \mathcal{F}^N,$
$$x \star y = \mathfrak{f}^{-1}\big( \mathfrak{f}(x) \cdot \mathfrak{f}(y) \big),$$
where $\cdot$ denotes scalar (element-wise) product.

There are algorithms for computing both discrete Fourier transform and inverse discrete Fourier transform
efficiently --- in $O(N \log N)$ time, provided that addition and multiplication
in $\mathcal{F}$ take $O(1)$ time. Any such algorithm is called a fast Fourier transform (FFT).

Although an $O(N \log N)$ FFT algorithm exists that works for arbitrary $N,$ even prime,
the most widely used FFT algorithm, the Cooley---Tukey algorithm,
works by rewriting a composite $N = N_1 N_2$ into $N_1$ smaller transforms of size $N_2.$
Thus, it works only for the values of $N$ which are highly composite;
particularly when $N$ is a power of two.

\subsection{Complex FFT using floating point}

The obvious approach to calculate a cyclic convolution over $\mathbb{Z}$ is to pick a field that
contains $\mathbb{Z}$ and has principal roots of unity of any order --- namely, $\mathbb{C},$ the
field of complex numbers.
A principal root of unity of order $n$ can then be expressed by the formula $e^{2 \pi i / n}.$
This is the core of the ``FFT multiplication'' algorithm usually taught in classes:
approximate $\mathbb{C}$ with a pair of double-precision floating point values, compute the
convolution and round the results back to integers.

The main drawback of this algorithm is the precision issues.
Formally, it is not even a sound algorithm for multiplication: provided that the precision of
your floating-point values is bounded, you can not multiply arbitrarily large numbers with it: at
some point, you are going to get a round-off error large enough to produce a wrong digit in the
answer.

In order to get a provably correct answer, you have to put less bits of information into the
floating-point values than otherwise possible. Because of that, in order to achieve performance
parity with multiplication algorithms that abstain from use of floating point, platform-specific
SIMD extensions are often used. Also, in order to get a provably correct result, you need to know
the precision of your floating-point types, as well as possible quirks of their implementation ---
i.e., depend on the platform.

\subsection{Number-theoretic transform}

For any prime number $p,$ there exists the finite field $\mathrm{F}_p$ of integers modulo
$p.$ The number-theoretic transform is defined as the discrete Fourier transform in
$\mathrm{F}_p,$ for some prime $p.$
Note that $\mathrm{F}_p$ contains a principal root of unity of order $n$ if and only if $n$ divides $(p-1).$

We will now discuss the application of number-theoretic transform to integer multiplication.

First note that, in practice, it is always possible to obtain a reasonable upper bound on the length
of numbers we would ever need to multiply.
Let us thus say we have chosen the maximum length of a multiplicand, $M \in \mathbb{N}.$

One approach is to pick a prime $p > (B-1)^2 M$ and perform computations in
$\mathrm{F}_p.$
Remember that the inputs to the convolution over $\mathbb{Z}$ we need to compute are in $[0; B-1]$
and outputs are in $[0; (B-1)^2 M],$ so doing calculations modulo $p$ would never result
in an ambiguity.
Note that, generally speaking, we would need to round the size of the convolution, $N,$ up to some
divisor of $(p-1)$ in order to guarantee the existence of principal root of unity of order $N.$
We thus need to choose $p$ such that $(p-1)$ is highly composite.

Another approach is to pick a number of pairwise distinct primes, $p_1, \ldots, p_n,$ such that
$$\prod\limits_{i=1}^{n} p_i > (B-1)^2 M.$$
Then, for each $i,$ compute the convolution vector in $\mathrm{F}_{p_i}$ and use
the Chinese remainder theorem to recover the actual answer.
In this case, we would need to round the size of the convolution, $N,$ up to some divisor of
$$\gcd(p_1 - 1, \ldots, p_n - 1).$$
We thus need to choose $p_1, \ldots, p_n$ such that this value is highly composite.

Note that the latter approach is just a generalization of the first one:
just assume that $\gcd(p-1) = p-1.$

Remember also that the ``high compositeness'' of $N,$ which must divide $\gcd(p_1 - 1, \ldots, p_n - 1),$ is required
for the Cooley-Tukey algorithm to work on a length-$N$ transform. What a coincidence!

\section{Choice of high-level algorithm}

Having chosen the number-theoretic transform, we now need to decide on the following:

\begin{itemize}
    \item what number of primes to use, and of what size (in machine words);

    \item what power of $10$ to choose as the base $B;$

    \item how to find primes with the qualities we desire.
\end{itemize}

\subsection{Choosing the size of primes in machine words}

All real-life hardware platforms have the notion of machine word; typically, we can natively add and
subtract machine words, and get the product of two machine words as a two-word value. The size of
the pointer is also typically limited to the machine word.
Suppose the machine word is $w$ bits long; define $\mu = 2^w.$

Let us fix, for every $\ell \ge 1,$ some representation in memory for elements of $\mathrm{F}_p,$
where $p$ is $\ell$ words long prime; this means that $\mu^{\ell - 1} < p < \mu^\ell.$
We then define the following functions:
\begin{itemize}
    \item $C_{\mathrm{Add}}(\ell),$ the cost of addition of two elements in $\mathrm{F}_p$ for $\ell$-word $p;$
    \item $C_{\mathrm{Sub}}(\ell),$ the cost of subtraction of two elements in $\mathrm{F}_p$ for $\ell$-word $p;$
    \item $C_{\mathrm{Mul}}(\ell),$ the cost of multiplication of two elements in $\mathrm{F}_p$ for $\ell$-word $p.$
\end{itemize}

Let us assume the following:
\begin{itemize}

    \item $\ell \cdot C_{\mathrm{Add}}(1) \le C_{\mathrm{Add}}(\ell).$
      Indeed, it should be impossible to add (subtract, compare) $\ell$-word numbers faster than
      doing $\ell$ single-word additions (subtractions, comparisons).
      Note that addition modulo $p$ is normally implemented as simple addition, comparison with
      $p$ (assuming no overflow) and conditional subtraction.

    \item $\ell \cdot C_{\mathrm{Sub}}(1) \le C_{\mathrm{Sub}}(\ell).$
      Similar to the above: subtraction modulo $p$ is normally implemented as simple subtraction,
      underflow test and conditional addition.

    \item $\ell \cdot C_{\mathrm{Mul}}(1) < C_{\mathrm{Mul}}(\ell)$ for $\ell > 1.$
      Assume $\ell$ is small enough that the optimal way to multiply two $\ell$-word numbers is the
      ``dumb'' quadratic algorithm.
      Even not considering the cost of reduction modulo $p,$
      the cost of ``raw'' multiplication of two $\ell$-word numbers into $2\ell$-word number is
      $\ell^2$ single-word multiplications with some additions, as opposed to just $\ell$
      single-word multiplications that the left-hand side of this inequality attempts to express.

\end{itemize}

Fix then a fast Fourier transform algorithm.
For a transform size of $N,$ it does $\Upsilon_{\mathrm{Add}}(N)$ additions, $\Upsilon_{\mathrm{Sub}}(N)$ subtractions, and $\Upsilon_{\mathrm{Mul}}(N)$
multiplications over $\mathrm{F}_p.$ We assume it does no divisions in $\mathrm{F}_p,$ which is a reasonable assumption if we have inverses in $\mathrm{F}_p$
to all possible values of $N$ pre-calculated.

We now want to compare the approach of using $\ell > 1$ pairwise distinct single-word primes against the approach of using a single $\ell$-word prime.
By our assumptions,
   $$\ell \big( C_{\mathrm{Add}}(1) \Upsilon_{\mathrm{Add}}(N) + C_{\mathrm{Sub}}(1) \Upsilon_{\mathrm{Sub}}(N) + C_{\mathrm{Mul}}(1) \Upsilon_{\mathrm{Mul}}(N) \big)$$
is less than
   $$C_{\mathrm{Add}}(\ell) \Upsilon_{\mathrm{Add}}(N) + C_{\mathrm{Sub}}(\ell) \Upsilon_{\mathrm{Sub}}(N) + C_{\mathrm{Mul}}(\ell) \Upsilon_{\mathrm{Mul}}(N),$$
if $\Upsilon_{\mathrm{Mul}}(N) > 0.$

This means that using $\ell$ distinct single-word primes is better;
the same argument can be invoked to show that using an ensemble of
primes of mixed word lengths is suboptimal compared to an ensemble of single-word primes.

Note that we do not consider the cost of answer recovery here,
because the transform part is $\Theta(N \log N)$
and the answer recovery part is $\Theta(N);$
and, in practice, the transform part dominates.

Neither does our analysis consider cache efficiency, which ought to be better for $\ell$ distinct
single-word primes.

\subsection{Choosing the number of primes}

Let us now choose the number of primes, $\ell.$
Remember $\mu = 2^w,$ where $w$ is the length of the machine word in bits.
Remember also that we want to pick $p_1, \ldots, p_\ell$ such that
$$\prod\limits_{i=1}^{\ell} p_i > (B-1)^2 M,$$
where $M$ is the maximum length of a multiplicand possible.

We will require $p_i < \mu/2$ for simpler implementation of addition modulo $p_i.$
If this does not hold, then, during addition of two numbers modulo $p_i,$ the raw sum may overflow
the machine word, and the implementation would need to check for two conditions instead of just
one: we would need to subtract $p_i$ (modulo $\mu$) from the result if either the overflow happened
or the result is greater or equal to $p_i.$

Having fixed $\mu = 2^w$ and $M,$ and assuming all $p_1, \ldots, p_\ell$ will be
approximately equal to $\mu/2,$ we can define the following function:
$$\lambda(\ell) = \max\{n \in \mathbb{N} : ({10}^n - 1)^2 < (\mu / 2)^\ell / M\}.$$
Then, once we choose $\ell,$ we calculate $\lambda(\ell)$ and, assuming it is positive, we can put
$B = {10}^{\lambda(\ell)}.$ If $\lambda(\ell) = 0,$ the chosen value of $\ell$ is too small; more
prime moduli are needed.

Note that we do not require $B \le \mu$ here; this may seem strange, but this
only impacts initialization and answer recovery stages, which are $\Theta(N),$ and may potentially
speed up the transform stage, which is $\Theta(N \log N).$

Let us calculate the values of $\lambda(\ell)$ for $\mu \in \{2^{32}, 2^{64}\},$ $M \in \{2^{15}, 2^{20}, 2^{25}\}$ and $\ell \in \{1,2,3,4\}.$

Note the meaning of $\lambda(\ell) / \ell$: if a multiplicand has $n$ digits in base ten, it will
have $\lceil n / \lambda(\ell) \rceil$ digits in base $B = 10^{\lambda(\ell)};$ but we will need to perform $\ell$
transforms: one for each prime modulo. The ratio $\lambda(\ell) / \ell,$ thus, is approximately the
``speedup factor'' over doing a single transform with $B = 10.$

First, for $\mu = 2^{64}$:

\medskip

\begin{tabular}{ lll }

\begin{tabular}{ |c|c|c| }
 \hline
 \multicolumn{3}{|c|}{ $M=2^{15}$ } \\
 \hline\hline
  $\ell$ & $\lambda(\ell)$ & $\lambda(\ell) / \ell$ \\
 \hline\hline
  $1$ & $7$ & $7.0000$ \\
 \hline
  $2$ & $16$ & $8.0000$ \\
 \hline
  $3$ & $26$ & $8.6667$ \\
 \hline
  $4$ & $35$ & $8.7500$ \\
 \hline
\end{tabular}
&
\begin{tabular}{ |c|c|c| }
 \hline
 \multicolumn{3}{|c|}{ $M=2^{20}$ } \\
 \hline\hline
  $\ell$ & $\lambda(\ell)$ & $\lambda(\ell) / \ell$ \\
 \hline\hline
  $1$ & $6$ & $6.0000$ \\
 \hline
  $2$ & $15$ & $7.5000$ \\
 \hline
  $3$ & $25$ & $8.3333$ \\
 \hline
  $4$ & $34$ & $8.5000$ \\
 \hline
\end{tabular}
&
\begin{tabular}{ |c|c|c| }
 \hline
 \multicolumn{3}{|c|}{ $M=2^{25}$ } \\
 \hline\hline
  $\ell$ & $\lambda(\ell)$ & $\lambda(\ell) / \ell$ \\
 \hline\hline
  $1$ & $5$ & $5.0000$ \\
 \hline
  $2$ & $15$ & $7.5000$ \\
 \hline
  $3$ & $24$ & $8.0000$ \\
 \hline
  $4$ & $34$ & $8.5000$ \\
 \hline
\end{tabular}
\end{tabular}

\medskip

Then, for $\mu = 2^{32}$:

\medskip

\begin{tabular}{ lll }

\begin{tabular}{ |c|c|c| }
 \hline
 \multicolumn{3}{|c|}{ $M=2^{15}$ } \\
 \hline\hline
  $\ell$ & $\lambda(\ell)$ & $\lambda(\ell) / \ell$ \\
 \hline\hline
  $1$ & $2$ & $2.0000$ \\
 \hline
  $2$ & $7$ & $3.5000$ \\
 \hline
  $3$ & $11$ & $3.6667$ \\
 \hline
  $4$ & $16$ & $4.0000$ \\
 \hline
\end{tabular}
&
\begin{tabular}{ |c|c|c| }
 \hline
 \multicolumn{3}{|c|}{ $M=2^{20}$ } \\
 \hline\hline
  $\ell$ & $\lambda(\ell)$ & $\lambda(\ell) / \ell$ \\
 \hline\hline
  $1$ & $1$ & $1.0000$ \\
 \hline
  $2$ & $6$ & $3.0000$ \\
 \hline
  $3$ & $10$ & $3.3333$ \\
 \hline
  $4$ & $15$ & $3.7500$ \\
 \hline
\end{tabular}
&
\begin{tabular}{ |c|c|c| }
 \hline
 \multicolumn{3}{|c|}{ $M=2^{25}$ } \\
 \hline\hline
  $\ell$ & $\lambda(\ell)$ & $\lambda(\ell) / \ell$ \\
 \hline\hline
  $1$ & $0$ & $0$ \\
 \hline
  $2$ & $5$ & $2.5000$ \\
 \hline
  $3$ & $10$ & $3.3333$ \\
 \hline
  $4$ & $14$ & $3.5000$ \\
 \hline
\end{tabular}
\end{tabular}

\medskip

First, note the phenomenon of diminishing returns: consider, for example, $\mu = 2^{64}, M=2^{20};$
the speedup factor of using $\ell=2$ over $\ell=1$ is $7.5 / 6 = 1.25,$
while the speedup factor of $\ell=3$ over $\ell=2$ is $8.3333 / 7.5 = 1.1111;$
and the speedup factor of $\ell=4$ over $\ell=3$ is $8.5 / 8.3333 = 1.02.$

At the same time, the larger $\ell$ is, the higher the costs of initialization and answer recovery
are.

Note also that, starting with $\ell=3,$ for any $2^{15} \le M \le 2^{25}$ and both $\mu=2^{32}$ and $\mu=2^{64},$
the value of $B = 10^{\lambda(\ell)}$ becomes greater than $\mu;$ this means that we can not represent a value
modulo $B$ with one machine word, which means more overhead (in both performance and complexity of the code) on
initialization and answer recovery.

We thus think it is wise to choose $\ell=2$ as a nice trade-off between the performance of the
transform, costs of initialization and answer recovery, and complexity of the code.

Note that the \textbf{mpdecimal} library uses $\ell = 3$ prime moduli with bases
$B=10^{19}$ on 64-bit systems and $B=10^{9}$ on 32-bit systems; the speedup factors are
$19/3=6.3333$ on 64-bit systems and $9/3=3$ on 32-bit systems.

\subsection{Choosing the base}

The definition of $\lambda(\ell)$ in the previous subsection gives us a way to calculate
$B = 10^{\lambda(\ell)}$ given the values of $\mu,$ $M,$ and $\ell.$
In practice, it is better to support multiple bases, each for its own range of transform lengths.
This eliminates the need for picking the single maximum transform length $M$ and slowing down
smaller transforms.

Our implementation supports bases
$\{ 10^{14}, 10^{15}, 10^{16}, 10^{17} \}$ on 64-bit systems; and bases
$\{ 10^{5}, 10^{6}, 10^{7} \}$ on 32-bit systems.

\subsection{Choosing the method of modular reduction}

The most computationally expensive low-level operation that is carried out during the
number-theoretic transform is modular multiplication. In order to multiply two elements of
$\mathrm{F}_p,$ it is not enough to simply compute the raw product of values in $[0; p-1]$:
we need to perform reduction modulo $p;$ although, as we will see, what it exactly means depends on
the representation of elements.

We discuss the methods of reduction now because one method requires prime moduli of special form;
thus, our choice may affect our strategy of searching primes.

\subsubsection{The naïve approach}

The naïve method of modular reduction is to use, where available, a hardware instruction
to divide two-words divident by single-word divisor into single-word quotient and single-word
remainder. Where not available, we would have to emulate such an instruction in software.

Note that such an instruction exists in x86 and x86-64.
As for software emulation, both GNU GCC and Clang compilers provide \verb!uint64_t! type on 32-bit
platforms, and \verb!unsigned __int128! type on 64-bit platforms,
with support for the division operation.

Unfortunately, this method is very slow. % TODO

\subsubsection{Barrett reduction}

A method intended to be faster, while not requiring a change in our representation of field
elements, is know as Barrett reduction. % TODO cite

It reduces $0 \le a < n^2$ modulo $n$ using some pre-computed value depending on $n.$
For a $k$-bit modulo $n,$ it internally (not counting the raw multiplication $xy = a$) performs:
\begin{itemize}
    \item one multiplication of $(k+1)$ bits by $(2k)$ bits into a $(3k)$-bit value;
    \item one multiplication of $k$ bits by $k$ bits, of which only the lowest $k$ bits are used.
\end{itemize}

\subsubsection{Montgomery reduction}

A method even faster for our purposes, is Montgomery reduction.

Remember we defined $w$ as the bit width of the machine word;
define then $R = 2^w \in \mathrm{F}_p.$ We assume $p > 2,$ so $R$ is non-zero in $\mathrm{F}_p.$

Then, the \textit{Montgomery representation} of $x \in \mathrm{F}_p$ is simply $Rx.$

If $Rx$ is the Montgomery representation of $x,$ and $Ry$ is the Montgomery representation of $y,$
then the Montgomery representation of $(x \pm y)$ is simply $(Rx \pm Ry)$ as $R(x \pm y) = Rx \pm Ry;$
it means that Montgomery representations can be added and subtracted as ordinary values modulo $p.$

The \textit{Montgomery reduction} is a function
$\redc \colon \mathrm{F}_p \times \mathrm{F}_p \to \mathrm{F}_p$
defined as follows:
$$\redc(x, y) = R^{-1} x y.$$
It is important because the Montgomery representation $R(x y)$ of the product $x y$ is
$\redc(Rx, Ry).$ Also, any $x \in \mathrm{F}_p$ can be converted into Montgomery
representation by invoking $\redc(x, R^2) = Rx;$ and out of Montgomery
representation by invoking $\redc(x, 1) = R^{-1}x.$

It is possible to compute $\redc$ efficiently; the algorithm internally
(not counting the raw $xy$ multiplication)
performs:
\begin{itemize}
    \item one multiplications of $w$ bits by $w$ bits into $(2w)$ bits;
    \item one multiplications of $w$ bits by $w$, of which only the lowest $w$ bits are used.
\end{itemize}
This is more efficient than the Barrett reduction if one does not consider the cost of conversion
into and out of Montgomery representation. In the case of number-theoretic transform, we can
convert everything into Montgomery representations at the beginning (this is $\Theta(N)$ time),
then perform the transform (this is $\Theta(N \log N)$ modular multiplications), and then
convert everything out of Montgomery representation (this is, again, $\Theta(N)$ time).

Better still, we can completely omit those conversions, instead mixing a factor into the final stage
of multiplication by $N^{-1}$ in the inverse transform; see section~\ref{section:trick} for details.

\subsubsection{Solinas primes}

The \textbf{mpdecimal} library uses yet another method of modular reduction on 64-bit systems;
it is based on the use of prime moduli of form $2^{64} - 2^{n} + 1.$
Generally, such primes are knows as Solinas primes, defined as primes of form
$f(2^m),$ where $f(x)$ is a low-degree polynomial with small integer coefficients.

The single round of reduction modulo $p = 2^{64} - 2^{n} + 1$ is then defined as follows:
$$r(2^{64} x_1 + x_0) = 2^{n} x_1 - x_1 + x_0,$$
where $x_0, x_1 \in \mathbb{N}$ and $x_0, x_1 < 2^{64}.$
After some small number of rounds (2---3 rounds for primes used in \textbf{mpdecimal}), the result
is guaranteed to be less than $2p,$ after which it can be reduced modulo $p$ with a single
conditional subtraction, just like with the Montgomery reduction.

We benchmarked the ``best case'' of reduction modulo Solinas prime against Montgomery reduction;
we used $2^{64} - 2^{32} + 1$ as the Solinas prime modulo; note the values of $64$ and $32$ are
better for hardware (at least, for x86-64) because division and multiplication can be done by
simply omitting (half-)words or assuming zero lower (half-)words, correspondingly, instead of
actually shifting the bits. This prime also requires at most 2 rounds of reduction.

We found that on modern system, reduction modulo Solinas prime is slower by $\approx$ 15\%;
the code can be found in the \verb!bench-solinas! subdirectory of our repository.

We think the most likely explanation for this finding is that, on modern x86-64 systems, hardware
multipliers are sufficiently performant to render special schemes for reduction modulo a
single-word $p$ useless, if Montgomery reduction can be used instead.

\subsection{Searching for primes}

Having chosen the Montgomery reduction, we are going search for primes of form
$$c \cdot 3 \cdot 2^n + 1.$$

The factor of $2^n$ here ensures that we will be able to perform transforms of length $2^m$ for any $m \le n;$
and the factor of $3$ is here to ``smooth the stairs'', allowing us to perform transforms of length
$3 \cdot 2^m$ for any $m \le n.$ Note that $3 \cdot 2^m = 1.5 \cdot 2^{m+1}$ is exactly the average
of $2^{m+1}$ and $2^{m+2}.$

Observe that, in order to guarantee the uniqueness of representation of a prime as
$c \cdot 3 \cdot 2^n + 1,$
we need to require $c$ to be odd --- otherwise a power of two can be factored out into $2^n.$

Define $n_{\mathrm{min}} = \lceil \log_2(M/3) \rceil,$ the lower bound for $n$ in the formula for
the desired form of our primes above.
We have, then, the following requirements for each $p_i$:
\begin{enumerate}
    \item $p_i$ is of form $c \cdot 3 \cdot 2^n + 1,$ where $c$ is odd and $n \ge n_{\mathrm{min}};$
    \item $p_i < \mu / 2.$
\end{enumerate}
Of all possible primes with those properties, we need to pick the $\ell$ largest.

This leads us to the following algorithm, presented here as Python code:

\begin{verbatim}
def find_for_n(n, p_max, max_nyield):
    """
    Yields at most 'max_nyield' largest primes of form
      p = c * 3 * 2**n + 1,
    such that p <= p_max and 'c' is odd.
    """

    # Calculate the largest factor 'f' such that:
    #   (1) f is odd;
    #   (2) f is divisible by 3;
    #   (3) f * 2**n + 1 <= p_max.

    f = (p_max - 1) >> n
    if f % 2 == 0:
        f -= 1
    while f % 3 != 0:
        f -= 2

    # Search for primes of form
    #   (f - 6*i) * 2**n + 1,
    # where i >= 0.

    while f > 0 and max_nyield:
        p = (f << n) + 1
        if is_prime(p):
            yield p
            max_nyield -= 1
        f -= 6

def find_primes(n_min, w, ell):
    """
    Returns list of 'ell' largest primes of form
      p = c * 3 * 2**n + 1,
    where:
      (1) n_min <= n <= (w - 3);
      (2) p < 2**(w - 1);
      (3) 'c' is odd.
    """

    result = []
    p_max = (1 << (w - 1)) - 1
    for n in range(n_min, w - 2):
        result.extend(find_for_n(n, p_max, ell))
    assert len(result) >= ell
    return sorted(result)[-ell:]
\end{verbatim}

Note that it uses the external function \verb|is_prime|, which must perform primality testing.
We use the Miller-Rabin primality test with the first 12 prime numbers as bases; it has been proven
in \cite{primebases} that, for values less than $2^{64},$ this is enough to guarantee correctness.

\section{The linearity trick}
\label{section:trick}

We are now returning to the Montgomery reduction.

Remember we defined $R = 2^w \in \mathrm{F}_p,$ where $w$ is the length of machine word in bits;
we assume $p > 2,$ so $R$ is non-zero in $\mathrm{F}_p.$

Remember also we defined the function
$\redc \colon \mathrm{F}_p \times \mathrm{F}_p \to \mathrm{F}_p$
in the following way:
$$\redc(x, y) = R^{-1} x y.$$

Consider the field $\widetilde{\mathrm{F}}_p,$ which is the same as $\mathrm{F}_p,$ but
having $\redc$ as the multiplication operation.
It is a field isomorphic to $\mathrm{F}_p,$ with isomorphism
$\varphi \colon \mathrm{F}_p \to \widetilde{\mathrm{F}}_p$ being $\varphi(x) = Rx.$

If we fix a principal root of unity $\xi \in \mathrm{F}_p$ of order $N,$
then $\widetilde{\xi} = R \xi$ will be a principal root of unity of order $N$ in
$\widetilde{\mathrm{F}}_p.$

We can consider, then,
$\widetilde{\mathfrak{f}}$ and $\widetilde{\mathfrak{f}}^{-1},$
the DFT and IDFT, correspondingly, for the field $\widetilde{\mathrm{F}}_p.$
We have the following identities, by the isomorphism, for any $x \in \mathrm{F}_p^N$:
\begin{eqnarray*}
    \mathfrak{f}(x)         &=& R^{-1} \, \widetilde{\mathfrak{f}}(R \, x); \\
    \mathfrak{f}^{-1}(x)    &=& R^{-1} \, \widetilde{\mathfrak{f}}^{-1}(R \, x).
\end{eqnarray*}

Rewrite with $x = R^{-1}y$:
\begin{eqnarray*}
    \mathfrak{f}(R^{-1}y)       &=& R^{-1} \, \widetilde{\mathfrak{f}}(y); \\
    \mathfrak{f}^{-1}(R^{-1}y)  &=& R^{-1} \, \widetilde{\mathfrak{f}}^{-1}(y).
\end{eqnarray*}

Rewrite the left-hand sides by the linearity of $\mathfrak{f}$ and $\mathfrak{f}^{-1}$:
\begin{eqnarray*}
    R^{-1} \mathfrak{f}(y)       &=& R^{-1} \, \widetilde{\mathfrak{f}}(y); \\
    R^{-1} \mathfrak{f}^{-1}(y)  &=& R^{-1} \, \widetilde{\mathfrak{f}}^{-1}(y).
\end{eqnarray*}

Dividing both sides by $R^{-1},$ we see that $\mathfrak{f}(y)$ and $\widetilde{\mathfrak{f}}(y)$
coincide for any $y = Rx;$ and so do $\mathfrak{f}^{-1}(y)$ and $\widetilde{\mathfrak{f}}^{-1}(y).$

Let us now consider the cyclic convolution $\widetilde{\star}$ for $\widetilde{\mathrm{F}}_p;$
we know that
$$x \, \widetilde{\star} \, y = \widetilde{\mathfrak{f}}^{-1}
\big( \thinspace \widetilde{\mathfrak{f}}(x) \, \widetilde{\otimes} \, \widetilde{\mathfrak{f}}(y) \big),$$
where $\widetilde{\otimes}$ denotes element-wise $\redc$
(which is, remember, the multiplication operation of $\widetilde{\mathrm{F}}_p.$)
Having proven that
$\widetilde{\mathfrak{f}} \equiv \mathfrak{f}$
and
$\widetilde{\mathfrak{f}}^{-1} \equiv \mathfrak{f}^{-1}$
above, we can rewrite it as follows:
$$x \, \widetilde{\star} \, y = R^{-1} \mathfrak{f}^{-1}\big( \mathfrak{f}(x) \cdot \mathfrak{f}(y) \big),$$
where $\cdot$ denotes element-wise product.
We can rewrite this as
$$x \, \widetilde{\star} \, y = R^{-1}(x \star y).$$

It means that, computing a cyclic convolution via DFT of both arguments, followed by element-wise
multiplication, followed by IDFT, but using
$\redc$ instead of regular multiplication during all those operation,
gives us an extra factor of $R^{-1}.$

Instead of converting both arguments into Montgomery representation
(remember this conversion is just multiplication by $R$), performing $\widetilde{\star}$ and
converting the result out of Montgomery representation (this is multiplication by $R^{-1}$),
we can convert only one of the arguments; indeed,
$$x \, \widetilde{\star} \, (Ry) = R^{-1}(x \star Ry) = x \star y.$$

Alternatively, we can multiply the result by $R,$ not touching any of the arguments at all:
$$R(x \, \widetilde{\star} \, y) = R R^{-1}(x \star y) = x \star y.$$

The latter approach is preferable for us: the Cooley-Tukey IDFT performs multiplication by
scalar $N^{-1}$ as a separate final stage; we can ``mix in'' the factor of $R$ into this stage
essentially for free!

Another advantage is that, it might be possible that we are computing the cyclic convolution of $x$
with itself: the arguments have not just identical values, but the same location in memory; so
multiplying one argument by $R$ automatically multiplies another by the same factor.

\bibliography{paper}{}
\bibliographystyle{plain}

\end{document}

### 2020-11-10 bench-solinas/bench.c

@@ -1,0 +1,93 @@
+#include <stdint.h>
+#include <stdio.h>
+#include <stdlib.h>
+
+#define ULIMB uint64_t
+#define SLIMB int64_t
+#define DOUBLE_ULIMB unsigned __int128
+#define LIMB_BITS 64
+#define FORCE_INLINE __attribute__((always_inline))
+
+static inline FORCE_INLINE
+DOUBLE_ULIMB full_prod(ULIMB x, ULIMB y)
+{
+    return ((DOUBLE_ULIMB) x) * y;
+}
+
+#if defined(USE_SOLINAS)
+
+#define MODULO 18446744069414584321ull
+static inline FORCE_INLINE
+ULIMB mul(ULIMB x, ULIMB y)
+{
+    DOUBLE_ULIMB t = full_prod(x, y);
+
+#define ROUND() \
+    do { \
+        ULIMB hi = t >> LIMB_BITS; \
+        ULIMB lo = t; \
+        t = (((DOUBLE_ULIMB) hi) << 32) - hi + lo; \
+    } while (0)
+
+    ROUND();
+    ROUND();
+
+#undef ROUND
+
+    // no adjustment
+    return t;
+}
+
+#elif defined(USE_MONT)
+
+#define MODULO    9203387313508319233ull
+#define MODULO_P1 9203387313508319231ull
+static inline FORCE_INLINE
+ULIMB mul(ULIMB x, ULIMB y)
+{
+    DOUBLE_ULIMB t = full_prod(x, y);
+    ULIMB m = t * MODULO_P1;
+    DOUBLE_ULIMB mp = full_prod(m, MODULO);
+    ULIMB r = (t + mp) >> LIMB_BITS;
+
+    // no adjustment
+    return r;
+}
+
+#else
+#   error "Please define either USE_SOLINAS or USE_MONT."
+#endif
+
+static int parse_int_or_die(const char *s)
+{
+    int r;
+    if (sscanf(s, "%d", &r) != 1) {
+        fprintf(stderr, "Cannot parse '%s' as integer.\n", s);
+        abort();
+    }
+    return r;
+}
+
+int main(int argc, char **argv)
+{
+    if (argc != 3) {
+        fprintf(stderr, "USAGE: %s seed nrepeat\n", argv[0]);
+        abort();
+    }
+    ULIMB x = parse_int_or_die(argv[1]);
+    int n = parse_int_or_die(argv[2]);
[…]

### 2020-11-10 bench.c

@@ -3,6 +3,7 @@
 #include <stdlib.h>
 #include <stdio.h>
 #include <string.h>
+#include <time.h>
 
 static void die_oom(void)
 {
@@ -53,26 +54,38 @@ static void write_arg(FFT_ULIMB *p, size_t n, size_t ndigits, unsigned base_log)
         p[i] = 0;
 }
 
-static size_t x_get_fft_size(unsigned method, size_t m)
+static size_t x_fft_size_aux(size_t init_factor, size_t m)
 {
-    size_t n;
-
-    switch (method) {
-    case 1:
-    case 6:
-        n = 1;
-        break;
-    case 4:
-        n = 3;
-        break;
-    default:
-        fprintf(stderr, "Unknown method: %u (expected either of: 1 4 6)\n", method);
-        abort();
-    }
-
+    size_t n = init_factor;
     while (n < m)
         n = x_mul_zu(n, 2);
     return n;
+}
+
+static size_t x_get_fft_size(unsigned *method, size_t m)
+{
+    switch (*method) {
+    case 1:
+    case 6:
+        return x_fft_size_aux(1, m);
+    case 4:
+        return x_fft_size_aux(3, m);
+    case 0:
+        {
+            size_t n_simple = x_fft_size_aux(1, m);
+            size_t n_fourstep = x_fft_size_aux(3, m);
+            if (n_simple < n_fourstep) {
+                *method = 1;
+                return n_simple;
+            } else {
+                *method = 4;
+                return n_fourstep;
+            }
+        }
+    default:
+        fprintf(stderr, "Unknown method: %u (expected either of: 0 1 4 6)\n", *method);
+        abort();
+    }
 }
 
 static int max_base_log(size_t ndigits)
@@ -105,21 +118,18 @@ int main(int argc, char **argv)
     size_t ndigits = x_parse_zu(argv[2]);
     size_t nrepeat = x_parse_zu(argv[3]);
 
-    size_t nres = x_mul_zu(ndigits, 2);
-
-    int base_log = max_base_log(nres);
+    int base_log = max_base_log(ndigits);
     if (base_log < 0) {
         fprintf(stderr, "Transform of such large size is not supported.\n");
         abort();
     }
 
[…]

### 2020-11-10 bench_mpdec.c

@@ -1,43 +1,64 @@
 #include <mpdecimal.h>
 #include <stdio.h>
-
-enum {
-	NDIGITS = 15000000,
-	NREPEAT = 5,
-};
+#include <stdlib.h>
+#include <time.h>
 
 static mpd_t *xalloc_mpd_t(void)
 {
-	mpd_t *r = mpd_qnew();
-	if (!r) {
-		fprintf(stderr, "Out of memory.\n");
-		abort();
-	}
-	return r;
+    mpd_t *r = mpd_qnew();
+    if (!r) {
+        fprintf(stderr, "Out of memory.\n");
+        abort();
+    }
+    return r;
 }
 
-int main()
+static int x_parse_int(const char *s)
 {
-	mpd_t *a = xalloc_mpd_t();
-	mpd_t *b = xalloc_mpd_t();
-	mpd_t *c = xalloc_mpd_t();
+    int r;
+    if (sscanf(s, "%d", &r) != 1) {
+        fprintf(stderr, "Cannot parse '%s' as integer.\n", s);
+        abort();
+    }
+    return r;
+}
 
-	mpd_context_t ctx;
-	mpd_maxcontext(&ctx);
+int main(int argc, char **argv)
+{
+    if (argc != 3) {
+        fprintf(stderr, "USAGE: bench_mpdec NDIGITS NREPEAT\n");
+        return 2;
+    }
+    int ndigits = x_parse_int(argv[1]);
+    int nrepeat = x_parse_int(argv[2]);
 
-	mpd_set_uint(a, 10, &ctx);
-	mpd_set_uint(b, NDIGITS, &ctx);
+    mpd_t *a = xalloc_mpd_t();
+    mpd_t *b = xalloc_mpd_t();
+    mpd_t *c = xalloc_mpd_t();
 
-	mpd_pow(a, a, b, &ctx);
+    mpd_context_t ctx;
+    mpd_maxcontext(&ctx);
 
-	mpd_sub_uint(a, a, 1, &ctx);
+    mpd_set_uint(a, 10, &ctx);
+    mpd_set_uint(b, ndigits, &ctx);
 
-	mpd_sub_uint(b, a, 1, &ctx);
+    mpd_pow(a, a, b, &ctx);
 
-	for (int i = 0; i < NREPEAT; ++i)
-		mpd_mul(c, a, b, &ctx);
+    mpd_sub_uint(a, a, 1, &ctx);
 
[…]

### 2020-11-10 fft.c

@@ -268,7 +268,7 @@ void mk_special_pow_table(FFT_ULIMB w, FFT_ULIMB *out, size_t n, const Field *fi
 
 // Assumes n is a power of two.
 static inline FFT_FORCE_INLINE
-void mul_poly(
+void convolve(
         FFT_ULIMB *a, FFT_ULIMB *b, size_t n,
         FFT_ULIMB *scratch,
         const Field *field)
@@ -278,18 +278,21 @@ void mul_poly(
     // make table of powers for root (in Montgomery representation)
     mk_special_pow_table(field->root_M, scratch, n, field);
 
+#if ! FFT_USE_LINEARITY_TRICK()
     // convert a into Montgomery representation
     for (size_t i = 0; i < n; ++i)
         a[i] = redc(a[i], field->mont_r2modp, field);
+#endif
 
     // forward transform on a
     dif_full(a, n, scratch, field);
 
     if (b != a) {
+#if ! FFT_USE_LINEARITY_TRICK()
         // convert b into Montgomery representation
         for (size_t i = 0; i < n; ++i)
             b[i] = redc(b[i], field->mont_r2modp, field);
-
+#endif
         // forward transform on b
         dif_full(b, n, scratch, field);
     }
@@ -304,8 +307,14 @@ void mul_poly(
     // inverse transform on a (without division by n)
     dit_full(a, n, scratch, field);
 
+#if ! FFT_USE_LINEARITY_TRICK()
     // divide by n and convert out of Montgomery representation
     FFT_ULIMB factor = field->inv_2[order];
+#else
+    // divide by n and account for Montgomery stuff
+    FFT_ULIMB r3 = redc(field->mont_r2modp, field->mont_r2modp, field);
+    FFT_ULIMB factor = redc(r3, field->inv_2[order], field);
+#endif
     for (size_t i = 0; i < n; ++i)
         a[i] = redc(a[i], factor, field);
 }
@@ -407,7 +416,7 @@ void fourstep_inverse(
 
 // Assumes n = 3*2^k.
 static inline FFT_FORCE_INLINE
-void mul_poly_with_4step(
+void convolve_fourstep(
         FFT_ULIMB *a, FFT_ULIMB *b, size_t n,
         FFT_ULIMB *scratch,
         const Field *field)
[…]

### 2020-11-10 fft.h

@@ -3,6 +3,8 @@
 #include <stdint.h>
 #include <stddef.h>
 
+#define FFT_USE_LINEARITY_TRICK() 0
+
 #if UINTPTR_MAX == 0xFFFFFFFFFFFFFFFFul && !defined(FFT_FORCE_32_BIT)
 #   define FFT_LIMB_BITS    64
 typedef uint64_t            FFT_ULIMB;

### 2020-11-10 gen/gen.py

@@ -2,78 +2,63 @@
 import argparse
 
 from isprime import is_prime
-from utils import egcd, canonical_mod, prim_root, mod_pow, modp_inv
+from utils import egcd, canonical_mod, prim_root, mod_pow, modp_inv, decompose_pow2
 import fastdiv
 
 
 class Platform:
-    def __init__(self, bitwidth, k_min, p_min, base_logs):
+    def __init__(self, bitwidth, k_min, base_logs):
         self.bitwidth = bitwidth
         self.k_min = k_min
-        self.p_min = p_min
         self.base_logs = base_logs
 
 
 PLATFORM_32 = Platform(
     bitwidth=32,
     k_min=25,
-    p_min=int(2e9),
     base_logs=[5, 6, 7])
 
 
 PLATFORM_64 = Platform(
     bitwidth=64,
-    k_min=47,
-    p_min=int(9.2e18),
+    k_min=35,
     base_logs=[14, 15, 16, 17])
 
 
-class Prime:
-    def __init__(self, c, k):
-        self.c = c
-        self.k = k
-        self.p = (c << k) + 1
-
-
-def generate_primes(p_min, p_max, k_min, k_max):
-
-    def _div_ceil(a, b):
-        q, r = divmod(a, b)
-        return q + (1 if r else 0)
-
-    def _round_up_c(n, k):
-        q = _div_ceil(_div_ceil(n, 1 << k), 3)
-        if q % 2 == 0:
-            q += 1
-        return 3 * q
-
-    def _round_down_c(n, k):
-        q = (n >> k) // 3
-        if q % 2 == 0:
-            q -= 1
-        return 3 * q
-
-    for k in range(k_max, k_min - 1, -1):
-        c_min = _round_up_c(p_min, k)
-        c_max = _round_down_c(p_max, k)
-        for c in range(c_max, c_min - 1, -6):
+def generate_primes_aux(k_min, k_max, p_max, npick):
+    assert npick > 0
+    for k in range(k_min, k_max + 1):
+        c_start = p_max >> k
+        if c_start % 2 == 0:
+            c_start -= 1
+        while c_start % 3 != 0:
+            c_start -= 2
+        nleft = npick
+        for c in range(c_start, -3, -6):
             p = (c << k) + 1
-            assert p >= p_min
-            assert p <= p_max
-            if not is_prime(p):
-                continue
-            yield Prime(c, k)
+            if is_prime(p):
+                yield p
[…]

### 2020-11-10 gen/isprime.py

@@ -1,7 +1,13 @@
 from utils import decompose_pow2, mod_pow
 
 
+# See: https://arxiv.org/pdf/1509.00864
+
+
 _BASES = [2, 3, 5, 7, 11, 13, 17, 19, 23, 29, 31, 37]
+
+
+_LIMIT = 1 << 64
 
 
 def _miller_rabin_round(n, a):
@@ -17,9 +23,15 @@ def _miller_rabin_round(n, a):
 
 
 def is_prime(n):
-    assert n > 40
-    assert n < (1 << 64)
-    assert (n & 1) != 0
+    if n >= _LIMIT:
+        raise ValueError('n is too large')
+
+    if n <= _BASES[-1]:
+        return n in _BASES
+
+    if (n & 1) == 0:
+        return False
+
     for base in _BASES:
         if not _miller_rabin_round(n, base):
             return False

### 2020-11-10 main.c

@@ -160,17 +160,16 @@ int main()
     char *b = read_number_str();
     size_t nb = strlen(b);
 
-    size_t nres = x_add_zu(na, nb);
-
-    int base_log = max_base_log(nres);
+    int base_log = max_base_log(na < nb ? na : nb);
     if (base_log < 0) {
         fprintf(stderr, "Transform of such large size is not supported.\n");
         abort();
     }
 
-    size_t nlimbs_m = div_ceil_zu(na > nb ? na : nb, base_log);
-    size_t nlimbs_r = div_ceil_zu(nres, base_log);
-    size_t n = x_calc_fft_size(2 * nlimbs_m);
+    size_t nlimbs_a = div_ceil_zu(na, base_log);
+    size_t nlimbs_b = div_ceil_zu(nb, base_log);
+    size_t nlimbs_r = x_add_zu(nlimbs_a, nlimbs_b);
+    size_t n = x_calc_fft_size(nlimbs_r);
 
     FFT_ULIMB *mem = x_calloc(sizeof(FFT_ULIMB), x_add_zu(4, x_mul_zu(n, 5)));
 

### 2020-11-12 paper/paper.tex
Update

@@
 \usepackage{mathrsfs}
-\usepackage{eufrak}
 \usepackage{cite}
@@
 Consider an aribtrary algebraic ring $\mathcal{R}.$
-The cyclic convolution $x \star y, \, \star \colon \mathcal{R}^n \times \mathcal{R}^n \to \mathcal{R}^n$ is defined as follows:
+The cyclic convolution $x \star y, \, (\_ \star \_) \colon \mathcal{R}^n \times \mathcal{R}^n \to \mathcal{R}^n$ is defined as follows:
 $$(x \star y)[k] = \sum\limits_{i=0}^{n - 1} x[i] y[(k - i) \mymod n].$$
@@
 
-We require that there exists a principal root of unity of order $N$ in $\mathcal{F}$:
-such $\xi \in \mathcal{F}$ that $\xi^N = 1$ and
-$$\sum\limits_{i=0}^{N-1} \xi^{ik} = 0$$
-for every $0 < k < N.$ We fix one such $\xi.$
+We require that there exists a primitive $N$-th root of unity in $\mathcal{F}$:
+such $\xi \in \mathcal{F}$ that $\xi^N = 1$ and $\xi^k \ne 1$ for every $0 < k < N.$
+We fix one such $\xi.$
 
@@
 
+Note that $\star$ is bilinear, which means that it is linear in both of its arguments:
+for any scalar $\alpha \in \mathcal{F}$ and any two vectors $x, y \in \mathcal{F}^N,$
+$$x \star (\alpha y) = (\alpha x) \star y = \alpha (x \star y).$$
+
 There are algorithms for computing both discrete Fourier transform and inverse discrete Fourier transform
@@
 The obvious approach to calculate a cyclic convolution over $\mathbb{Z}$ is to pick a field that
-contains $\mathbb{Z}$ and has principal roots of unity of any order --- namely, $\mathbb{C},$ the
+contains $\mathbb{Z}$ and has primitive $n$-th roots of unity for any $n$ --- namely, $\mathbb{C},$ the
 field of complex numbers.
-A principal root of unity of order $n$ can then be expressed by the formula $e^{2 \pi i / n}.$
+A primitive $n$-th root of unity can then be expressed by the formula $e^{2 \pi i / n}.$
 This is the core of the ``FFT multiplication'' algorithm usually taught in classes:
@@
 $\mathrm{F}_p,$ for some prime $p.$
-Note that $\mathrm{F}_p$ contains a principal root of unity of order $n$ if and only if $n$ divides $(p-1).$
+Note that $\mathrm{F}_p$ contains a primitive $n$-th root of unity if and only if $n$ divides $(p-1).$
 
@@
 Note that, generally speaking, we would need to round the size of the convolution, $N,$ up to some
-divisor of $(p-1)$ in order to guarantee the existence of principal root of unity of order $N.$
+divisor of $(p-1)$ in order to guarantee the existence of primitive $N$-th root of unity.
 We thus need to choose $p$ such that $(p-1)$ is highly composite.
 
-Another approach is to pick a number of pairwise distinct primes, $p_1, \ldots, p_n,$ such that
+Another approach is to pick $n$ pairwise distinct primes, $p_1, \ldots, p_n,$ such that
 $$\prod\limits_{i=1}^{n} p_i > (B-1)^2 M.$$
@@
 
-The naïve method of modular reduction is to use, where available, a hardware instruction
+The naïve method of modular reduction is to use, where it is available, a hardware instruction
 to divide two-words divident by single-word divisor into single-word quotient and single-word
@@
 
-It reduces $0 \le a < n^2$ modulo $n$ using some pre-computed value depending on $n.$
+It reduces $0 \le a < n^2$ modulo $n$ using some pre-computed value that depends on $n.$
 For a $k$-bit modulo $n,$ it internally (not counting the raw multiplication $xy = a$) performs:
@@
 
-We benchmarked the ``best case'' of reduction modulo Solinas prime against Montgomery reduction;
-we used $2^{64} - 2^{32} + 1$ as the Solinas prime modulo; note the values of $64$ and $32$ are
-better for hardware (at least, for x86-64) because division and multiplication can be done by
-simply omitting (half-)words or assuming zero lower (half-)words, correspondingly, instead of
+We benchmarked the ``best case'' of reduction modulo $2^{N_1}-2^{N_2}+1$ against Montgomery
+reduction: we used $2^{64} - 2^{32} + 1$ as the prime modulo; note the values of $64$ and
+$32$ are better for hardware (at least, for x86-64) because division and multiplication can be done
+by simply omitting (half-)words or assuming zero lower (half-)words, correspondingly, instead of
 actually shifting the bits. This prime also requires at most 2 rounds of reduction.
 
-We found that on modern system, reduction modulo Solinas prime is slower by $\approx$ 15\%;
+We found that on a modern x86-64 system, reduction with two rounds of $r$ is slower by
+$\approx$15\% compared to Montgomery reduction;
 the code can be found in the \verb!bench-solinas! subdirectory of our repository.
@@
 multipliers are sufficiently performant to render special schemes for reduction modulo a
-single-word $p$ useless, if Montgomery reduction can be used instead.
+single-word $p=2^{N_1}-2^{N_2}+1$ useless, if Montgomery reduction can be used instead.
 
@@
 
-If we fix a principal root of unity $\xi \in \mathrm{F}_p$ of order $N,$
-then $\widetilde{\xi} = R \xi$ will be a principal root of unity of order $N$ in
-$\widetilde{\mathrm{F}}_p.$
+If we fix a primitive $N$-th root of unity $\xi \in \mathrm{F}_p,$
+then $\widetilde{\xi} = R \xi$ will be a primitive $N$-th root of unity
+in $\widetilde{\mathrm{F}}_p.$
 
@@
 where $\cdot$ denotes element-wise product.
-We can rewrite this as
+This, in part, can be rewritten as
 $$x \, \widetilde{\star} \, y = R^{-1}(x \star y).$$
@@
 multiplication, followed by IDFT, but using
-$\redc$ instead of regular multiplication during all those operation,
+$\redc$ instead of regular multiplication during all of those operations,
 gives us an extra factor of $R^{-1}.$
@@
 converting the result out of Montgomery representation (this is multiplication by $R^{-1}$),
-we can convert only one of the arguments; indeed,
+we can just convert one of the arguments; indeed,
 $$x \, \widetilde{\star} \, (Ry) = R^{-1}(x \star Ry) = x \star y.$$
@@
 
-Another advantage is that, it might be possible that we are computing the cyclic convolution of $x$
-with itself: the arguments have not just identical values, but the same location in memory; so
-multiplying one argument by $R$ automatically multiplies another by the same factor.
+Note that multiplication by $R$ is $\redc$ with $R^2;$ so if the Montgomery representation of
+the factor $N^{-1}$ was $\psi = RN^{-1},$
+then our new factor will be
+$$\psi' = \redc(\psi, R^2) = R^2 N^{-1}.$$
+Then, $\forall x \in \mathrm{F}_p \colon \redc(x, \psi') = R N^{-1} x,$ exactly what we need.
+
+Another practical advantage of modification of the result as opposed to one of the arguments is
+that, it might be possible that we are computing the cyclic convolution of the sequence with
+\textit{itself}: the sequences have not \textit{just} identical values, but
+\textit{the same location in memory}; so multiplying one by $R$ automatically multiplies another.
+Of course, whether such a situation is allowed depends on the implementation; but we do support
+such a use case.
+
+\section{The matrix Fourier algorithms}
+
+We now describe the four-step and six-step algorithms for DFT/IDFT, which are matrix Fourier
+algorithms, meaning that they interpret sequences of length $N = N_1 N_2$ as $N_1 \times N_2$ matrices.
+
+We assume that rows, as opposed to columns, occupy contiguous ranges in underlying sequences;
+in other words, we assume row-major order:
+$$
+\langle 1, 2, 3, 4, 5, 6, 7, 8 \rangle \longleftrightarrow
+\begin{bmatrix}
+1 & 2 & 3 & 4 \\
+5 & 6 & 7 & 8
+\end{bmatrix}.
+$$
+
+\subsection{The four-step algorithm}
+
+For the purposes of this subsection:
+\begin{itemize}
+
+    \item $[N_1 \times N_2]$ means ``interpreting the sequence as a matrix with $N_1$ rows and $N_2$ columns''.
+
+    \item If a matrix has $N_1$ rows and $N_2$ columns, then its element with
+        row index $i,$ where $0 \le i < N_1,$
+        and column index $j,$ where $0 \le j < N_2,$ is said to be indexed $\langle i, j \rangle.$
+
+\end{itemize}
+
+Fix the length of transform $N.$
+Fix also $\xi_N,$ a primitive $N$-th root of unity in the underlying field.
+
+DFT of length $N = RC$ can be computed as follows:
+\begin{enumerate}
+    \item $[R \times C]$, perform (length $R$) DFT on each column with $\xi = (\xi_N)^C.$
+    \item $[R \times C]$, multiply each element indexed $\langle i,j \rangle$ by $(\xi_N)^{ij}.$
+    \item $[R \times C]$, perform (length $C$) DFT on each row with $\xi = (\xi_N)^R.$
+    \item $[R \times C]$, transpose the matrix.
+\end{enumerate}
+
+IDFT of length $N = RC$ can be computed as follows:
+\begin{enumerate}
+    \item $[C \times R],$ transpose the matrix.
+    \item $[R \times C],$ perform (length $C$) IDFT on each row with $\xi = (\xi_N)^R.$
+    \item $[R \times C],$ multiply each element indexed $\langle i,j \rangle$ by $(\xi_N)^{-ij}.$
+    \item $[R \times C],$ perform (length $R$) IDFT on each column with $\xi = (\xi_N)^C.$
+\end{enumerate}
+
+\subsection{The six-step algorithm}
+
+For the purposes of this subsection:
+\begin{itemize}
+
+    \item $[N_1 \times N_2]$ means ``interpreting the sequence as a matrix with $N_1$ rows and $N_2$ columns''.
+
+    \item If a matrix has $N_1$ rows and $N_2$ columns, then its element with
+        row index $i,$ where $0 \le i < N_1,$
+        and column index $j,$ where $0 \le j < N_2,$ is said to be indexed $\langle i, j \rangle.$
+
+\end{itemize}
+
+Fix the length of transform $N.$
+Fix also $\xi_N,$ a primitive $N$-th root of unity in the underlying field.
+
+DFT of length $N = RC$ can be computed as follows:
+\begin{enumerate}
+    \item $[R \times C],$ transpose the matrix.
+    \item $[C \times R],$ perform (length $R$) DFT on each row with $\xi = (\xi_N)^C.$
+    \item $[C \times R],$ transpose the matrix.
+    \item $[R \times C],$ multiply each element indexed $\langle i,j \rangle$ by $(\xi_N)^{ij}.$
+    \item $[R \times C],$ perform (length $C$) DFT on each row with $\xi = (\xi_N)^R.$
+    \item $[R \times C],$ transpose the matrix.
+\end{enumerate}
+
+IDFT of length $N = RC$ can be computed as follows:
+\begin{enumerate}
+    \item $[C \times R],$ transpose the matrix.
+    \item $[R \times C],$ perform (length $C$) IDFT on each row with $\xi = (\xi_N)^R.$
+    \item $[R \times C],$ multiply each element indexed $\langle i,j \rangle$ by $(\xi_N)^{-ij}.$
+    \item $[R \times C],$ transpose the matrix.
+    \item $[C \times R],$ perform (length $R$) IDFT on each row with $\xi = (\xi_N)^C.$
+    \item $[C \times R],$ transpose the matrix.
+\end{enumerate}
+
+\subsection{Discussion}
+
+If $\xi_N$ is a primitive $N$-th root of unity in a field $\mathcal{F},$
+and $N = N_1 N_2,$ then $\xi_{N_1} = (\xi_N)^{N_2}$ is a primitive $N_1$-th root of unity in
+$\mathcal{F}.$
+This means that we can indeed perform DFTs/IDFTs on rows/columns with the ``custom'' primitive
+roots specified in the descriptions of the algorithms.
+
+The four-step algorithm is just a restatement, in terms of a $N_1 \times N_2$ matrix, of the
+Cooley-Tukey algorithm that, in its general form, re-expresses a transform of length $N = N_1 N_2$
+in terms of smaller transforms of lengths $N_1$ and $N_2.$
+Thus, it can be used when it is desirable to perform a transform of length $N$ which is not a
+power of two, so that the decimation-in-time (DIT) or decimation-in-frequency (DIF) variants of the
+Cooley-Tukey algorithm can not be employed directly.
+
+The six-step algorithm has the advantage that, aside from matrix transpositions, it only accesses
+memory in strides of $R$ or $C.$ Thus, it might be faster in settings of hierarchial memory,
+including external-memory transforms and modern systems with multiple levels of cache.
+
+Remember our definition of the cyclic convolution in terms of DFT and IDFT:
+$$x \star y = \mathfrak{f}^{-1}\big( \mathfrak{f}(x) \cdot \mathfrak{f}(y) \big),$$
+where $\cdot$ denotes scalar (element-wise) product.
+
+Note that, for computation of $\star,$ the specific order in which the direct transform
+$\mathfrak{f}$ produces its outputs, and the order in which the inverse transform
+$\mathfrak{f}^{-1}$ expects its inputs to be in, does not matter; we only need these two orders
+to match.
+Indeed, $\cdot$ is element-wise, $(x \cdot y)[i] = x[i] y[i],$
+so, for any permutation $\pi$ we have $$\pi^{-1}(\pi(x) \cdot \pi(y)) = x \cdot y.$$
+This means that, for the purpose of computing the cyclic convolution, in both four-step and six-step
+algorithms, the final transposition step in DFT and the initial transposition step in IDFT, can be
+simultaneously omitted. Note this does not mean you can omit just one of them but not another!
 

### 2020-11-12 bench_gmp.c

@@ -1,0 +1,52 @@
+#include <gmp.h>
+#include <stdio.h>
+#include <stdlib.h>
+#include <time.h>
+
+static int x_parse_int(const char *s)
+{
+    int r;
+    if (sscanf(s, "%d", &r) != 1) {
+        fprintf(stderr, "Cannot parse '%s' as integer.\n", s);
+        abort();
+    }
+    return r;
+}
+
+int main(int argc, char **argv)
+{
+    if (argc != 4) {
+        fprintf(stderr, "USAGE: bench_gmp BASE NDIGITS NREPEAT\n");
+        return 2;
+    }
+    int base = x_parse_int(argv[1]);
+    int ndigits = x_parse_int(argv[2]);
+    int nrepeat = x_parse_int(argv[3]);
+
+    mpz_t x;
+    mpz_init(x);
+    mpz_ui_pow_ui(x, base, ndigits);
+    mpz_sub_ui(x, x, 1);
+
+    mpz_t y;
+    mpz_init_set(y, x);
+    mpz_sub_ui(y, y, 1);
+
+    mpz_t z;
+    mpz_init(z);
+
+    fprintf(stderr, "base=%d, ndigits=%d, nrepeat=%d\n", base, ndigits, nrepeat);
+    clock_t t = clock();
+
+    for (int i = 0; i < nrepeat; ++i) {
+        mpz_mul(z, x, y);
+    }
+
+    t = clock() - t;
+    printf("%.4f\n", ((double) t) / CLOCKS_PER_SEC);
+
+    mpz_clear(x);
+    mpz_clear(y);
+    mpz_clear(z);
+    return 0;
+}

### 2020-11-12 fft.c

@@ -70,9 +70,14 @@ FFT_DOUBLE_ULIMB full_prod(FFT_ULIMB x, FFT_ULIMB y)
 static inline FFT_FORCE_INLINE
 FFT_ULIMB modp_adjust(FFT_ULIMB x, const Field *field)
 {
+#if FFT_USE_BUILTIN_UNPREDICTABLE()
+    FFT_SLIMB d = x - field->p;
+    return __builtin_unpredictable(d < 0) ? x : d;
+#else
     FFT_SLIMB d = x - field->p;
     FFT_SLIMB cf = d >> (FFT_LIMB_BITS - 1);
     return d + (field->p & cf);
+#endif
 }
 
 // Performs addition modulo field->p.
@@ -100,9 +105,15 @@ FFT_ULIMB modp_add3(FFT_ULIMB x, FFT_ULIMB y, FFT_ULIMB z, const Field *field)
 static inline FFT_FORCE_INLINE
 FFT_ULIMB modp_sub(FFT_ULIMB x, FFT_ULIMB y, const Field *field)
 {
+#if FFT_USE_BUILTIN_UNPREDICTABLE()
+    FFT_SLIMB d = x - y;
+    FFT_SLIMB s = __builtin_unpredictable(d < 0) ? field->p : 0;
+    return d + s;
+#else
     FFT_SLIMB d = x - y;
     FFT_SLIMB cf = d >> (FFT_LIMB_BITS - 1);
     return d + (field->p & cf);
+#endif
 }
 
 // Performs Montgomery reduction on (x * y).

### 2020-11-12 fft.h

@@ -3,7 +3,9 @@
 #include <stdint.h>
 #include <stddef.h>
 
-#define FFT_USE_LINEARITY_TRICK() 0
+#define FFT_USE_LINEARITY_TRICK() 1
+
+#define FFT_USE_BUILTIN_UNPREDICTABLE() 1
 
 #if UINTPTR_MAX == 0xFFFFFFFFFFFFFFFFul && !defined(FFT_FORCE_32_BIT)
 #   define FFT_LIMB_BITS    64

### 2020-11-12 fft_compdep.h

@@ -2,9 +2,9 @@
 
 #include <stddef.h>
 
-# define FFT_UNUSED          __attribute__((unused))
-# define FFT_FORCE_INLINE    __attribute__((always_inline))
-# define FFT_UNREACHABLE()   __builtin_unreachable()
+#define FFT_UNUSED          __attribute__((unused))
+#define FFT_FORCE_INLINE    __attribute__((always_inline))
+#define FFT_UNREACHABLE()   __builtin_unreachable()
 
 static inline FFT_UNUSED FFT_FORCE_INLINE
 unsigned fft_zu_counttz(size_t x)

### 2020-11-12 gen/gen.py

@@ -13,6 +13,9 @@ class Platform:
         self.base_logs = base_logs
 
 
+BASE = 10
+
+
 PLATFORM_32 = Platform(
     bitwidth=32,
     k_min=25,
@@ -56,7 +59,7 @@ def find_primes_for_platform(platform, nprimes):
     return result
 
 
-def get_k(p):
+def extract_k_from_prime(p):
     k, _ = decompose_pow2(p - 1)
     return k
 
@@ -71,7 +74,7 @@ def print_consts_for_platform(platform, prime1, prime2):
         mont_p1 = canonical_mod(-neg_mont_p1, MONT_R)
         x = prim_root(prime)
 
-        k = get_k(prime)
+        k = extract_k_from_prime(prime)
 
         print(f'static const FFT_ULIMB FFT_ARRAY_{number}[] =')
         print('{')
@@ -114,7 +117,7 @@ def print_thresholds_for_platform(platform, prime1, prime2):
     print('// Auto-generated; do not edit.')
 
     product = prime1 * prime2
-    min_k = min(get_k(p) for p in [prime1, prime2])
+    min_k = min(extract_k_from_prime(p) for p in [prime1, prime2])
     max_transform_len = 3 << min_k
 
     print('const FFT_Threshold FFT_THRESHOLDS[] =')
@@ -127,8 +130,8 @@ def print_thresholds_for_platform(platform, prime1, prime2):
         return f'UINT64_C({n})'
 
     for base_log in sorted(platform.base_logs, reverse=True):
-        base = 10 ** base_log
-        max_n = (product - 1) // ((base - 1) ** 2)
+        radix = BASE ** base_log
+        max_n = (product - 1) // ((radix - 1) ** 2)
         max_n = min(max_n, max_transform_len // 2)
         ndigits = max_n * base_log
         print(' {NDIGITS(%s), %s},' % (make_const(ndigits), base_log))
@@ -157,10 +160,8 @@ def print_fastdiv_for_platform(platform, prime1, prime2):
         return lo, hi
 
     for base_log in platform.base_logs:
-        base = 10 ** base_log
-        c = fastdiv.fastdiv(d=base, N=platform.bitwidth * 2)
-        if not isinstance(c, fastdiv.CaseEasy):
-            raise NotImplementedError()
+        radix = BASE ** base_log
+        c = fastdiv.fastdiv(d=radix, N=platform.bitwidth * 2)
 
[…]

### 2020-11-14 paper/paper.tex
Update

@@
 \usepackage[english]{babel}
-\usepackage{mathtools}
 \usepackage{amssymb}
 \usepackage{amsmath}
-\usepackage{amsthm}
 \usepackage{mathrsfs}
 \usepackage{cite}
+\usepackage[ruled]{algorithm2e}
 
@@
 
+\section{Notation}
+
+The following notation is employed throughout this document:
+$$\langle x_0, \ldots, x_{n-1} \rangle [i] = x_i.$$
+In order words, square brackets mean zero-based tuple indexing.
+
+By $\mathbb{N}$ we mean $\{0, 1, 2, \ldots\}.$
+
+A ring is a set $\mathcal{R}$ equipped with two binary operations, $+$ and $\cdot,$ such that
+$\langle \mathcal{R}, + \rangle$ is an abelian group,
+$\cdot$ is associative and there an identity element with respect to $\cdot$ in $\mathcal{R},$
+and $\cdot$ is both left distributive and right distributive with respect to $+.$
+
 \section{Overview of the discrete Fourier transform}
@@
 
-The following notation is employed throughout this section:
-$$\langle x_0, \ldots, x_{n-1} \rangle [i] = x_i.$$
-
-By $\mathbb{N}$ we mean $\{0, 1, 2, \ldots\}.$
-
 \subsection{Cyclic convolution}
@@
 $$\myllangle x_0, \ldots, x_{n-1} \myrrangle = \sum\limits_{i=0}^{n-1} x_i B^i,$$
-where $x_i \in \mathbb{N}, \, x_i < B.$ In order words, square brackets mean zero-based tuple indexing.
+where $x_i \in \mathbb{N}, \, x_i < B.$
 
@@
 We will pretend natural numbers are in $\mathcal{F}$ by adopting the following identity:
-$$n = \underbrace{\overline{1} + \overline{1} + \cdots + \overline{1}}_\text{$n$ times},$$
-where $\overline{1} \in \mathcal{F}$ is the multiplicative identity in $\mathcal{F};$ the degenerate case of this identity is $0 = \overline{0},$
-where $\overline{0} \in \mathcal{F}$ is the additive identity in $\mathcal{F}.$
+$$n = \underbrace{\widehat{1} + \widehat{1} + \cdots + \widehat{1}}_\text{$n$ times},$$
+where $\widehat{1} \in \mathcal{F}$ is the multiplicative identity in $\mathcal{F};$ the degenerate case of this identity is $0 = \widehat{0},$
+where $\widehat{0} \in \mathcal{F}$ is the additive identity in $\mathcal{F}.$
 
@@
 Although an $O(N \log N)$ FFT algorithm exists that works for arbitrary $N,$ even prime,
-the most widely used FFT algorithm, the Cooley---Tukey algorithm,
-works by rewriting a composite $N = N_1 N_2$ into $N_1$ smaller transforms of size $N_2.$
+the most widely used FFT algorithm, the Cooley---Tukey algorithm, works by re-writing a transform
+of composite length $N = N_1 N_2$ into smaller transforms of lengths $N_1$ and $N_2.$
 Thus, it works only for the values of $N$ which are highly composite;
-particularly when $N$ is a power of two.
+particularly, if $N$ is a power of two, its variants named
+decimation in time (DIT) and decimation in frequency (DIF), can be used.
 
@@
 
-The main drawback of this algorithm is the precision issues.
+The main drawback of this algorithm is precision issues.
 Formally, it is not even a sound algorithm for multiplication: provided that the precision of
@@
 A method intended to be faster, while not requiring a change in our representation of field
-elements, is know as Barrett reduction. % TODO cite
+elements, is known as Barrett reduction. % TODO cite
 
@@
 We found that on a modern x86-64 system, reduction with two rounds of $r$ is slower by
-$\approx$15\% compared to Montgomery reduction;
+$\approx$ 15\% compared to Montgomery reduction;
 the code can be found in the \verb!bench-solinas! subdirectory of our repository.
@@
 
-This leads us to the following algorithm, presented here as Python code:
-
-\begin{verbatim}
-def find_for_n(n, p_max, max_nyield):
-    """
-    Yields at most 'max_nyield' largest primes of form
-      p = c * 3 * 2**n + 1,
-    such that p <= p_max and 'c' is odd.
-    """
-
-    # Calculate the largest factor 'f' such that:
-    #   (1) f is odd;
-    #   (2) f is divisible by 3;
-    #   (3) f * 2**n + 1 <= p_max.
-
-    f = (p_max - 1) >> n
-    if f % 2 == 0:
-        f -= 1
-    while f % 3 != 0:
-        f -= 2
-
-    # Search for primes of form
-    #   (f - 6*i) * 2**n + 1,
-    # where i >= 0.
-
-    while f > 0 and max_nyield:
-        p = (f << n) + 1
-        if is_prime(p):
-            yield p
-            max_nyield -= 1
-        f -= 6
-
-def find_primes(n_min, w, ell):
-    """
-    Returns list of 'ell' largest primes of form
-      p = c * 3 * 2**n + 1,
-    where:
-      (1) n_min <= n <= (w - 3);
-      (2) p < 2**(w - 1);
-      (3) 'c' is odd.
-    """
-
-    result = []
-    p_max = (1 << (w - 1)) - 1
-    for n in range(n_min, w - 2):
-        result.extend(find_for_n(n, p_max, ell))
-    assert len(result) >= ell
-    return sorted(result)[-ell:]
-\end{verbatim}
-
-Note that it uses the external function \verb|is_prime|, which must perform primality testing.
+This leads us to the following algorithm:
+
+\begin{procedure}[H]
+\caption{FindPrimesForN($n$, $p_{\max}, \ell$)}
+\SetKw{KwAnd}{and}
+\DontPrintSemicolon
+\KwIn{$n, p_{\max}, \ell.$}
+\KwOut{Set of $\ell$ largest, or less if the total number is less than $\ell,$ primes of form $p = c \cdot 3 \cdot 2^n + 1,$ where $p \le p_{\max}$ and $c$ is odd.}
+\Begin{
+    $r \longleftarrow \varnothing$\;
+    $\psi \longleftarrow (p_{\max} - 1) \mydiv 2^n $\;
+    \If{$\psi \mymod 2 = 0$}{
+        $\psi \longleftarrow \psi - 1$\;
+    }
+    \While{$\psi \mymod 3 \ne 0$}{
+        $\psi \longleftarrow \psi - 2$\;
+    }
+    \While{$\psi > 0$ \KwAnd $|r| < \ell$}{
+        $p \longleftarrow \psi \cdot 2^n + 1$\;
+        \If{$p$ is prime}{
+            $r \longleftarrow r \cup \{p\}$\;
+        }
+        $\psi \longleftarrow \psi - 6$\;
+    }
+    \Return{$r$}\;
+}
+\end{procedure}
+
+\begin{procedure}[H]
+\caption{FindPrimes($w$, $n_{\min}$, $\ell$)}
+\SetKw{Error}{error}
+\DontPrintSemicolon
+\KwIn{$w, n_{\min}, \ell.$}
+\KwOut{Set of $\ell$ largest primes of form $p = c \cdot 3 \cdot 2^n + 1,$ where $p < 2^{w-1},$ $n \ge n_{\min},$ and $c$ is odd.}
+\Begin{
+    $r \longleftarrow \varnothing$\;
+    $p_{\max} \longleftarrow 2^{w-1}-1$\;
+    \For{$n \longleftarrow n_{\min}$ \KwTo $w-2$}{
+        $r \longleftarrow r \cup \mathrm{FindPrimesForN}(n, p_{\max}, \ell)$\;
+    }
+    \eIf{$|r| < \ell$}{
+        \Error{cannot find $\ell$ primes with required properties}\;
+    }{
+        \Return{$\ell$ largest values of $r$}\;
+    }
+}
+\end{procedure}
+
+Note that it uses an unspecified algorithm to perform primality testing.
 We use the Miller-Rabin primality test with the first 12 prime numbers as bases; it has been proven
 in \cite{primebases} that, for values less than $2^{64},$ this is enough to guarantee correctness.
+
+\section{Cooley-Tukey optimizations}
+
+Before discussing optimizations,
+we need to define the decimation in time (DIT) and decimation in frequency (DIF) variants of the
+Cooley-Tukey FFT algorithm.
+
+Both are specializations of Cooley-Tukey for a power-of-two length, and both include the step of
+applying the bit-reversal permutation, which we introduce below, to the array being transformed,
+either at the very beginning or in the very end.
+
+Both algorithms take two parameters, an array $A$ of field elements and a field element
+$\omega = \xi^{\pm 1},$
+and return an array $A'$ of length $|A'| = |A|$
+such that $$A'[k] = \sum\limits_{i=0}^{|A| - 1} A[i] \omega^{ik}.$$
+If we express this transform as a function $\tau(A,\omega),$ then DFT and IDFT can be expressed in
+terms of $\tau$ as follows:
+\begin{eqnarray*}
+\mathfrak{f}(x) &=& \tau(x, \xi); \\
+\mathfrak{f}^{-1}(x) &=& N^{-1} \tau(x, \xi^{-1}),
+\end{eqnarray*}
+where $N$ is the length of the transform.
+
+\subsection{Bit-reversal permutation}
+
+Bit-reversal permutation is defined for sequences of length $2^n$ as follows: element with
+zero-based index $k$ is exchanged with element with zero-based index $\mathbf{rev}_n(k),$ where
+$\mathbf{rev}_n(k)$ is defined as the unique number in $[0; 2^n - 1]$ whose length-$n$ binary
+representation equals to the reversed length-$n$ binary representation of $k.$
+
+We now give the definition of the procedure that performs bit-reversal permutation.
+
+\begin{procedure}[H]
+\caption{BitRevPermute($A$)}
+\DontPrintSemicolon
+\KwIn{Array $A$, $|A| = 2^n.$}
+\KwOut{The result of applying the bit-reversal permutation to $A.$}
+\Begin{
+    $n \longleftarrow \log_2 |A|$\;
+    \For{$k \longleftarrow 0$ \KwTo $2^n - 1$}{
+        $\langle b_0, \ldots, b_{n-1} \rangle \longleftarrow$ bits of $k;$ $b_i \in \{0,1\}$ and $\sum_{i=0}^{n-1} b_i 2^i = k$\;
+        $k' \longleftarrow \sum_{i=0}^{n-1} b_{n-i-1} 2^i$\;
+        \If{$k < k'$}{
+            {exchange $A[k]$ with $A[k']$}\;
+        }
+    }
+    \Return{$A$}\;
+}
+\end{procedure}
+
+\subsection{Decimation in time}
+
+We now define the decimation in time (DIT) algorithm.
+Note that it performs bit-reversal permutation as the first step.
+$\widehat{1}$ denotes the multiplicative identity of the underlying field.
+
+\begin{procedure}[H]
+\caption{DecimationInTime($A$, $\omega$)}
+\SetKw{KwBy}{by}
+\DontPrintSemicolon
+\KwIn{
+    Array $A$ of field elements, $|A| = 2^n$.
+    Field element $\omega = \xi^{\pm 1},$ where $\xi$ is a primitive $2^n$-th root of unity.
+}
+\KwOut{
+    Array $A'$ such that $|A'| = |A|, A'[k] = \sum_{i=0}^{|A| - 1} A[i] \omega^{ik}.$
+}
+\Begin{
+    $A \longleftarrow \mathrm{BitRevPermute}(A)$\;
+    \For{$s \longleftarrow 1$ \KwTo $\log_2 |A|$}{
+        $m \longleftarrow 2^s$\;
+        $\omega_m \longleftarrow \omega^{|A| / m}$\;
+        \For{$k = 0$ \KwTo $|A|-1$ \KwBy $m$}{
+            $\varphi \longleftarrow \widehat{1}$\;
+            \For{$j = 0$ \KwTo $m/2-1$}{
+                $u \longleftarrow A[k + j]$\;
+                $v \longleftarrow \varphi \cdot A[k + j + m/2]$\;
+                $A[k + j] \longleftarrow u + v$\;
+                $A[k + j + m/2] \longleftarrow u - v$\;
+                $\varphi \longleftarrow \varphi \cdot \omega_m$\;
+            }
+        }
+    }
+    \Return{$A$}\;
+}
+\end{procedure}
+
+The four-line transform on $A[k+j]$ and $A[k+j+m/2]$ is known as ``butterfly'', or,
+more specifically, this one is the Cooley-Tukey butterfly; it maps
+$\langle x,y \rangle \mapsto \langle x', y' \rangle$ as follows:
+\begin{eqnarray*}
+x' &=& x + \varphi \cdot y; \\
+y' &=& x - \varphi \cdot y.
+\end{eqnarray*}
+
+\subsection{Decimation in frequency}
+
+We now define the decimation in frequency (DIF) algorithm.
+Note that it performs bit-reversal permutation as the last step.
+As in the previous subsection, $\widehat{1}$ denotes the multiplicative identity of the underlying field.
+
+\begin{procedure}[H]
+\caption{DecimationInFrequency($A$, $\omega$)}
+\SetKw{KwBy}{by}
+\SetKw{KwDownTo}{downto}
+\DontPrintSemicolon
+\KwIn{
+    Array $A$ of field elements, $|A| = 2^n$.
+    Field element $\omega = \xi^{\pm 1},$ where $\xi$ is a primitive $2^n$-th root of unity.
+}
+\KwOut{
+    Array $A'$ such that $|A'| = |A|, A'[k] = \sum_{i=0}^{|A| - 1} A[i] \omega^{ik}.$
+}
+\Begin{
+    \For{$s \longleftarrow \log_2 |A|$ \KwDownTo $1$}{
+        $m \longleftarrow 2^s$\;
+        $\omega_m \longleftarrow \omega^{|A| / m}$\;
+        \For{$k = 0$ \KwTo $|A|-1$ \KwBy $m$}{
+            $\varphi \longleftarrow \widehat{1}$\;
+            \For{$j = 0$ \KwTo $m/2-1$}{
+                $u \longleftarrow A[k + j]$\;
+                $v \longleftarrow A[k + j + m/2]$\;
+                $A[k + j] \longleftarrow u + v$\;
+                $A[k + j + m/2] \longleftarrow \varphi \cdot (u - v)$\;
+                $\varphi \longleftarrow \varphi \cdot \omega_m$\;
+            }
+        }
+    }
+    $A \longleftarrow \mathrm{BitRevPermute}(A)$\;
+    \Return{$A$}\;
+}
+\end{procedure}
+
+The four-line transform on $A[k+j]$ and $A[k+j+m/2]$ is known as the Gentleman-Sande butterfly; it maps
+$\langle x,y \rangle \mapsto \langle x', y' \rangle$ as follows:
+\begin{eqnarray*}
+x' &=& x + y; \\
+y' &=& \varphi \cdot (x - y).
+\end{eqnarray*}
+
+\subsection{Omitting the bit-reversal permutation}
+
+Remember we expressed the cyclic convolution in terms of DFT and IDFT as follows:
+$$x \star y = \mathfrak{f}^{-1}\big( \mathfrak{f}(x) \cdot \mathfrak{f}(y) \big),$$
+where $\cdot$ denotes scalar (element-wise) product.
+
+Note that, for the computation of $(\_ \star \_),$ the specific order in which the direct transform
+$\mathfrak{f}$ produces its outputs, and the order in which the inverse transform
+$\mathfrak{f}^{-1}$ expects its inputs to be in, do not matter; we only need these two orders
+to match.
+Indeed, $(\_ \cdot \_)$ is element-wise, $(x \cdot y)[i] = x[i] y[i],$
+so, for any permutation $\pi$ we have $$\pi^{-1}(\pi(x) \cdot \pi(y)) = x \cdot y.$$
+
+This means that, for the purpose of computing the cyclic convolution, we can use DIF for the direct
+transform and DIT for the inverse transform, and simultaneously omit the final permutation step in
+DIF and the initial permutation step in DIT.
+But note this does not mean just one of them can be omitted but not the other!
+
+\subsection{Pre-calculating the table of root powers}
+
+\subsection{Separate factor-1 butterfly}
 
@@
 
-The four-step algorithm is just a restatement, in terms of a $N_1 \times N_2$ matrix, of the
+The four-step algorithm is just a restatement, in terms of $N_1 \times N_2$ matrix, of the
 Cooley-Tukey algorithm that, in its general form, re-expresses a transform of length $N = N_1 N_2$
 in terms of smaller transforms of lengths $N_1$ and $N_2.$
-Thus, it can be used when it is desirable to perform a transform of length $N$ which is not a
-power of two, so that the decimation-in-time (DIT) or decimation-in-frequency (DIF) variants of the
+Thus, it can be used when it is desirable to perform a transform of a length that is not a
+power of two, so that the decimation in time (DIT) or decimation in frequency (DIF) variants of the
 Cooley-Tukey algorithm can not be employed directly.
@@
 The six-step algorithm has the advantage that, aside from matrix transpositions, it only accesses
-memory in strides of $R$ or $C.$ Thus, it might be faster in settings of hierarchial memory,
+memory in strides of $R$ or $C.$ Thus, it might be faster in settings of hierarchical memory,
 including external-memory transforms and modern systems with multiple levels of cache.
 
-Remember our definition of the cyclic convolution in terms of DFT and IDFT:
-$$x \star y = \mathfrak{f}^{-1}\big( \mathfrak{f}(x) \cdot \mathfrak{f}(y) \big),$$
-where $\cdot$ denotes scalar (element-wise) product.
-
-Note that, for computation of $\star,$ the specific order in which the direct transform
-$\mathfrak{f}$ produces its outputs, and the order in which the inverse transform
-$\mathfrak{f}^{-1}$ expects its inputs to be in, does not matter; we only need these two orders
-to match.
-Indeed, $\cdot$ is element-wise, $(x \cdot y)[i] = x[i] y[i],$
-so, for any permutation $\pi$ we have $$\pi^{-1}(\pi(x) \cdot \pi(y)) = x \cdot y.$$
-This means that, for the purpose of computing the cyclic convolution, in both four-step and six-step
-algorithms, the final transposition step in DFT and the initial transposition step in IDFT, can be
-simultaneously omitted. Note this does not mean you can omit just one of them but not another!
+\subsection{Omitting matrix transpositions}
+
+\subsection{Algorithm for matrix transposition}
+
+\subsection{Optimizing the four-step algorithm for $N=3 \cdot 2^k$}
+
+\subsection{Optimizing the six-step algorithm for $N=2^k$}
+
+\section{Low-level details}
+
+\subsection{Modular addition and subtraction}
+
+\subsection{Answer recovery}
+
+\subsection{Power table}
+
+\subsection{Shenanigans with Montgomery reduction}
+
+Remember the algorithm for computing $\redc^\ast(t) = R^{-1} t$
+in $\mathrm{F}_p,$
+assumes $$0 \le t < p \mu,$$ where $\mu = 2^w$ if the machine word is $w$ bits long.
+We can then compute $\redc(x,y),$ where $0 \le x,y < p$, as $\redc^\ast(xy).$
+
+Remember also that we require $p < \mu/2.$
 

### 2020-11-15 paper/paper.tex
Update

@@
 \subsubsection{Montgomery reduction}
+\label{section:montgomery_intro}
 
@@
 
-It is possible to compute $\redc$ efficiently; the algorithm internally
-(not counting the raw $xy$ multiplication)
+It is possible to compute $\redc$ efficiently. We now give the definition of
+the procedure $\mathrm{MontgomeryReduce}(a).$
+If we represent field elements as values in $[0; p-1],$ then
+$\redc(x,y)$ can be computed as $\mathrm{MontgomeryReduce}(xy).$
+
+\begin{procedure}[H]
+\DontPrintSemicolon
+\caption{MontgomeryReduce($a$)}
+\KwData{
+    Integer $\mu.$\;
+    Integer $p$ coprime with $\mu.$\;
+    Integer $p' \in [0; \mu-1]$ such that $p \cdot p' \equiv -1 \, (\mymod \mu).$
+}
+\KwIn{
+    Integer $a \in [0; \mu p - 1].$
+}
+\KwOut{
+    Integer $b \in [0; p-1]$ such that $b \equiv a \mu^{-1} \, (\mymod p).$
+}
+\Begin{
+    $m \longleftarrow ((a \mymod \mu) \cdot p') \mymod \mu$\;
+    $r \longleftarrow (a + m \cdot p) \mydiv \mu$
+        \tcc*[r]{this division is exact} % apparently no need to use \; after \tcc
+    \eIf{$r \ge p$}{
+        \Return{$r - p$}\;
+    }{
+        \Return{$r$}\;
+    }
+}
+\end{procedure}
+
+The value of $p'$ can be pre-calculated with extended Euclidean algorithm for GCD;
+we have $\gcd(p, \mu) = 1,$ so $\exists A, B \in \mathbb{Z} \colon A p + B \mu = 1.$
+This means $A p \equiv 1 \, (\mymod \mu)$ and $-Ap \equiv -1 \, (\mymod \mu),$ so
+$$p' = (-Ap) \mymod \mu.$$
+
+The procedure $\mathrm{MontgomeryReduce(a)}$ internally
+(not counting the raw multiplication $xy = a$)
 performs:
@@
 \subsection{Omitting the bit-reversal permutation}
+\label{section:omitting1}
 
@@
 
+Define $N = \log_2 |A|.$
+
+Note that, in both DIT and DIF, we perform the same computation over and over again when doing
+$$\varphi \longleftarrow \varphi \cdot \omega_m,$$
+where $\omega_m = \omega^{2^{N - s}}, s = \log_2 m.$
+
+One way to improve that is to pre-calculate powers of $\omega$ into a table $t,$ so that
+$t[i] = \omega^i, 0 \le i < |A|/2.$
+Then, given $j$ and $m=2^s,$ we can express the factor $\varphi$ as $t[j \cdot 2^{N - s}].$
+Note that the multiplication by a power of two can be replaced with a bit shift.
+
+Another way is to pre-calculate a separate table for each ``granularity'' of root powers:
+define $$T_s[i] = t[i \cdot 2^{N-s}] = (\omega^{2^{N-s}})^i, 0 \le i < 2^{s-1}$$
+for each $s, 1 \le s \le N.$ Note $T_N = t$ and $\sum\limits_{s=1}^{N} |T_s| = \sum\limits_{s=1}^{N} 2^{s-1} = 2^N-1.$
+Now, given $j$ and $m=2^s,$ we can express $\varphi$ as $T_s[j].$
+
+Let us write $T_s$ for all $s$ into a single array $T_\ast$
+of length $2^N-1 = |A|-1$ so that
+$$T_s[i] = T_\ast[(2^0 + 2^1 + \cdots + 2^{s-2}) + i] = T_\ast[2^{s-1} - 1 + i].$$
+Then, given $j$ and $m=2^s,$ the factor $\varphi$ can be expressed as
+$T_\ast[\Delta_s + j],$ where $\Delta_s = 2^{s-1}-1 = m/2 - 1.$
+
+At first sight, there is not much difference between indexing
+$T_\ast[C_1 + j]$ and $t[j \cdot 2^{C_2}]$
+in the innermost loop,
+where $C_1$ and $C_2$ denote loop-invariant expressions that can be calculated once before the loop.
+But on x86 and x86-64, there is difference:
+these platforms can encode SIB (scale-index-base) directly in the instruction that dereferences the
+pointer, but only for constant scales of 1, 2, 4 and 8 bytes.
+This means that, if our field element is 8, 16, 32 or 64 bits long, we can dereference $p[j],$
+where $p$ points to $T_\ast[C_1],$ in one instruction; but dereferencing
+$t[j \cdot 2^{C_2}],$ where $C_2$ can be anything, requires two instructions: first for the bit
+shift and the second for dereferencing.
+
 \subsection{Separate factor-1 butterfly}
+
+Note that, in both DIT and DIF, the factor $\varphi$ for the step $j=0$ is $\widehat{1},$
+the multiplicative identity of the field.
+In this case, both butterflies degenerate into
+$\langle x, y \rangle \mapsto \langle x+y, x-y \rangle.$
+
+We can peel the first iteration off the loop to avoid redundant multiplication by $\widehat{1}.$
 
@@
 
+Remember in section~\ref{section:omitting1} we established that, with
+$x \cdot y$ denoting scalar (element-wise) product between $x,y \in \mathcal{F}^N,$
+for any permutation $\pi$ we have $$\pi^{-1}(\pi(x) \cdot \pi(y)) = x \cdot y.$$
+
+We then used this property to get rid of bit-reversal permutations in DIT and DIF.
+It turns out that, for both four-step and six-step algorithms,
+the same property can be employed to get rid of the
+final transposition step in DFT and the initial transposition step in IDFT simultaneously.
+
+For the four-step algorithm, this means no bit-reversal permutation or matrix transposition is ever
+needed: we can also simultaneously omit the bit-reversal permutation steps in DIF and DIT
+if we use DIF for the direct transforms in definition of four-step DFT, and DIT for inverse
+transforms in the definition of four-step IDFT.
+
+For the six-step algorithm, unfortunately, things are not so simple.
+We can simultaneously omit the final transposition step (step 6) in the direct transform
+and the initial transposition step (step 1) in the inverse transform; and can use DIF without
+bit-reversal permutations for step 5 of the direct transform and DIT without bit-reversal
+permutations for step 2 of the inverse transform; but we still have two matrix transposition steps
+per transform, and no matter whether we use DIT or DIF for step 2 of the direct transform
+and step 5 of the inverse transform, we need to perform bit-reversal permutation.
+
 \subsection{Algorithm for matrix transposition}
@@
 
-Remember the algorithm for computing $\redc^\ast(t) = R^{-1} t$
-in $\mathrm{F}_p,$
-assumes $$0 \le t < p \mu,$$ where $\mu = 2^w$ if the machine word is $w$ bits long.
-We can then compute $\redc(x,y),$ where $0 \le x,y < p$, as $\redc^\ast(xy).$
-
-Remember also that we require $p < \mu/2.$
+Remember the procedure $\mathrm{MontgomeryReduce}(a)$ that we defined in section~\ref{section:montgomery_intro}.
 

### 2020-11-18 paper/paper.tex
Update

@@
 The cyclic convolution $x \star y, \, (\_ \star \_) \colon \mathcal{R}^n \times \mathcal{R}^n \to \mathcal{R}^n$ is defined as follows:
-$$(x \star y)[k] = \sum\limits_{i=0}^{n - 1} x[i] y[(k - i) \mymod n].$$
+\begin{equation} \label{eq:star}
+(x \star y)[k] = \sum\limits_{i=0}^{n - 1} x[i] y[(k - i) \mymod n].
+\end{equation}
 
@@
 
-Note that $c_i \le (B-1)^2 \min\{n, m\}$ as $x_i < B, \, y_i < B,$
-and, because of our zero padding, the summation in the definition of the cyclic convolution
-can have at most $\min\{n,m\}$ non-zero summands.
+Note that
+\begin{equation} \label{eq:ci_bound}
+c_i \le (B-1)^2 \min\{n, m\}
+\end{equation}
+as both $x_i$ and $y_i$ are less that $B$ and, because of our zero padding, the summation
+in~\eqref{eq:star} can have at most $\min\{n,m\}$ non-zero summands.
 
 The representation of $z$ in base $B$ can then be recovered by the following iterative process:
-\begin{eqnarray*}
-\sigma_0 &=& 0; \\
-z_i &=& (\sigma_i + c_i) \mymod B; \\
-\sigma_{i+1} &=& (\sigma_i + c_i) \mydiv B.
-\end{eqnarray*}
+\begin{subequations} \label{eq:sigmai_and_zi}
+\begin{align}
+\sigma_0 &= 0; \\
+z_i &= (\sigma_i + c_i) \mymod B; \\
+\sigma_{i+1} &= (\sigma_i + c_i) \mydiv B.
+\end{align}
+\end{subequations}
 
@@
 
-Define $L = (B-1)^2 \min\{n,m\}$ and remember $c_i \le L.$
+Define $L = (B-1)^2 \min\{n,m\}$ and remember that in~\eqref{eq:ci_bound}, we established that
+$c_i \le L.$
 Define also the following sequence:
@@
 We have
-$$\sigma_i \le a_i \le a^\prime_i < \lim\limits_{i \to \infty} a'_i = \frac{L}{B - 1} = (B-1)\min\{n, m\}.$$
+\begin{equation} \label{eq:sigmai_bound}
+\sigma_i \le a_i \le a^\prime_i < \lim\limits_{i \to \infty} a'_i = \frac{L}{B - 1} = (B-1)\min\{n, m\}.
+\end{equation}
 
@@
 Discrete Fourier transform $\mathfrak{f} \colon \mathcal{F}^N \to \mathcal{F}^N$ is then defined as follows:
-$$\mathfrak{f}(x)[k] = \sum\limits_{i=0}^{N-1} x[i] \xi^{ik}.$$
+\begin{equation} \label{eq:dft}
+\mathfrak{f}(x)[k] = \sum\limits_{i=0}^{N-1} x[i] \xi^{ik}.
+\end{equation}
 
@@
 up to the negated exponent of $\xi$ and multiplication by $N^{-1}$:
-$$\mathfrak{f}^{-1}(x)[k] = N^{-1} \sum\limits_{i=0}^{N-1} x[i] \xi^{-ik}.$$
+\begin{equation} \label{eq:idft}
+\mathfrak{f}^{-1}(x)[k] = N^{-1} \sum\limits_{i=0}^{N-1} x[i] \xi^{-ik}.
+\end{equation}
 
@@
 $\alpha \in \mathcal{F}$ and $x \in \mathcal{F}^N,$
-$\mathfrak{f}(\alpha x) = \alpha \mathfrak{f}(x)$
-and
-$\mathfrak{f}^{-1}(\alpha x) = \alpha \mathfrak{f}^{-1}(x).$ Here, the product of a scalar and a vector
+\begin{subequations} \label{eq:transform_linearity}
+\begin{align}
+\mathfrak{f}(\alpha x) &= \alpha \mathfrak{f}(x); \\
+\mathfrak{f}^{-1}(\alpha x) &= \alpha \mathfrak{f}^{-1}(x).
+\end{align}
+\end{subequations}
+Above, the product of a scalar and a vector
 $\alpha v = \alpha \langle v_0, \ldots, v_{n-1} \rangle$ denotes, as usual,
@@
 The convolution theorem says that, for any two vectors $x, y \in \mathcal{F}^N,$
-$$x \star y = \mathfrak{f}^{-1}\big( \mathfrak{f}(x) \cdot \mathfrak{f}(y) \big),$$
+\begin{equation} \label{eq:conv_theorem}
+x \star y = \mathfrak{f}^{-1}\big( \mathfrak{f}(x) \cdot \mathfrak{f}(y) \big),
+\end{equation}
 where $\cdot$ denotes scalar (element-wise) product.
@@
 for any scalar $\alpha \in \mathcal{F}$ and any two vectors $x, y \in \mathcal{F}^N,$
-$$x \star (\alpha y) = (\alpha x) \star y = \alpha (x \star y).$$
+\begin{equation} \label{eq:star_linearity}
+x \star (\alpha y) = (\alpha x) \star y = \alpha (x \star y).
+\end{equation}
 
@@
 Another approach is to pick $n$ pairwise distinct primes, $p_1, \ldots, p_n,$ such that
-$$\prod\limits_{i=1}^{n} p_i > (B-1)^2 M.$$
+\begin{equation} \label{eq:product_inequality}
+\prod\limits_{i=1}^{n} p_i > (B-1)^2 M.
+\end{equation}
 Then, for each $i,$ compute the convolution vector in $\mathrm{F}_{p_i}$ and use
@@
 Let us now choose the number of primes, $\ell.$
-Remember $\mu = 2^w,$ where $w$ is the length of the machine word in bits.
-Remember also that we want to pick $p_1, \ldots, p_\ell$ such that
+Remember we defined $\mu = 2^w,$ where $w$ is the length of the machine word in bits.
+Remember also that by~\eqref{eq:product_inequality},
+we want to pick $p_1, \ldots, p_\ell$ such that
 $$\prod\limits_{i=1}^{\ell} p_i > (B-1)^2 M,$$
@@
 
-We will require $p_i < \mu/2$ for simpler implementation of addition modulo $p_i.$
+We will require
+\begin{equation} \label{eq:pi_bound}
+p_i < \mu/2
+\end{equation}
+for simpler implementation of addition modulo $p_i.$
 If this does not hold, then, during addition of two numbers modulo $p_i,$ the raw sum may overflow
@@
 approximately equal to $\mu/2,$ we can define the following function:
-$$\lambda(\ell) = \max\{n \in \mathbb{N} : ({10}^n - 1)^2 < (\mu / 2)^\ell / M\}.$$
+\begin{equation} \label{eq:lambda}
+\lambda(\ell) = \max\{n \in \mathbb{N} : ({10}^n - 1)^2 < (\mu / 2)^\ell / M\}.
+\end{equation}
 Then, once we choose $\ell,$ we calculate $\lambda(\ell)$ and, assuming it is positive, we can put
@@
 
-The definition of $\lambda(\ell)$ in the previous subsection gives us a way to calculate
+Definition~\eqref{eq:lambda} gives us a way to calculate
 $B = 10^{\lambda(\ell)}$ given the values of $\mu,$ $M,$ and $\ell.$
@@
 Remember we defined $w$ as the bit width of the machine word;
-define then $R = 2^w \in \mathrm{F}_p.$ We assume $p > 2,$ so $R$ is non-zero in $\mathrm{F}_p.$
+define then
+\begin{equation} \label{eq:mont_r}
+R = 2^w \in \mathrm{F}_p.
+\end{equation}
+We assume $p > 2,$ so $R$ is non-zero in $\mathrm{F}_p.$
 
@@
 defined as follows:
-$$\redc(x, y) = R^{-1} x y.$$
+\begin{equation} \label{eq:mont_redc}
+\redc(x, y) = R^{-1} x y.
+\end{equation}
 It is important because the Montgomery representation $R(x y)$ of the product $x y$ is
@@
     Integer $p$ coprime with $\mu.$\;
-    Integer $p' \in [0; \mu-1]$ such that $p \cdot p' \equiv -1 \, (\mymod \mu).$
+    Integer $p' \in [0; \mu-1]$ such that $p \cdot p' \equiv -1 \, (\mymod \, \mu).$
 }
@@
 \KwOut{
-    Integer $b \in [0; p-1]$ such that $b \equiv a \mu^{-1} \, (\mymod p).$
+    Integer $b \in [0; p-1]$ such that $b \equiv a \mu^{-1} \, (\mymod \, p).$
 }
@@
 
-The value of $p'$ can be pre-calculated with extended Euclidean algorithm for GCD;
-we have $\gcd(p, \mu) = 1,$ so $\exists A, B \in \mathbb{Z} \colon A p + B \mu = 1.$
-This means $A p \equiv 1 \, (\mymod \mu)$ and $-Ap \equiv -1 \, (\mymod \mu),$ so
-$$p' = (-Ap) \mymod \mu.$$
+For the pre-calculated data, put $\mu = 2^w,$ $p > 2$ is our prime number, and
+the value of $p'$ can be calculated with the extended Euclidean algorithm for GCD;
+we have $\gcd(p, \mu) = 1,$ so $\exists c_1, c_2 \in \mathbb{Z} \colon c_1 p + c_2 \mu = 1.$
+This means $c_1 p \equiv 1 \, (\mymod \, \mu)$ and $-c_1 p \equiv -1 \, (\mymod \, \mu),$ so
+$$p' = (-c_1 p) \mymod \mu.$$
 
@@
 Having chosen the Montgomery reduction, we are going search for primes of form
-$$c \cdot 3 \cdot 2^n + 1.$$
+\begin{equation} \label{eq:pformula}
+c \cdot 3 \cdot 2^n + 1.
+\end{equation}
 
@@
 
-Observe that, in order to guarantee the uniqueness of representation of a prime as
-$c \cdot 3 \cdot 2^n + 1,$
+Observe that, in order to guarantee the uniqueness of representation of a prime as~\eqref{eq:pformula},
 we need to require $c$ to be odd --- otherwise a power of two can be factored out into $2^n.$
 
-Define $n_{\mathrm{min}} = \lceil \log_2(M/3) \rceil,$ the lower bound for $n$ in the formula for
-the desired form of our primes above.
+Define $n_{\mathrm{min}} = \lceil \log_2(M/3) \rceil,$ the lower bound for $n$ in~\eqref{eq:pformula}.
 We have, then, the following requirements for each $p_i$:
@@
 such that $$A'[k] = \sum\limits_{i=0}^{|A| - 1} A[i] \omega^{ik}.$$
-If we express this transform as a function $\tau(A,\omega),$ then DFT and IDFT can be expressed in
-terms of $\tau$ as follows:
-\begin{eqnarray*}
-\mathfrak{f}(x) &=& \tau(x, \xi); \\
-\mathfrak{f}^{-1}(x) &=& N^{-1} \tau(x, \xi^{-1}),
-\end{eqnarray*}
-where $N$ is the length of the transform.
+If we express this transform as a function $\tau(A,\omega),$ then DFT and IDFT of length $N$ can be
+expressed in terms of $\tau$ as follows:
+\begin{subequations}
+\begin{align}
+\mathfrak{f}(x) &= \tau(x, \xi); \\
+\mathfrak{f}^{-1}(x) &= N^{-1} \tau(x, \xi^{-1}).
+\end{align}
+\end{subequations}
 
@@
 $\langle x,y \rangle \mapsto \langle x', y' \rangle$ as follows:
-\begin{eqnarray*}
-x' &=& x + \varphi \cdot y; \\
-y' &=& x - \varphi \cdot y.
-\end{eqnarray*}
+\begin{subequations}
+\begin{align}
+x' &= x + \varphi \cdot y; \\
+y' &= x - \varphi \cdot y.
+\end{align}
+\end{subequations}
 
@@
 $\langle x,y \rangle \mapsto \langle x', y' \rangle$ as follows:
-\begin{eqnarray*}
-x' &=& x + y; \\
-y' &=& \varphi \cdot (x - y).
-\end{eqnarray*}
+\begin{subequations}
+\begin{align}
+x' &= x + y; \\
+y' &= \varphi \cdot (x - y).
+\end{align}
+\end{subequations}
 
@@
 
-Remember we expressed the cyclic convolution in terms of DFT and IDFT as follows:
-$$x \star y = \mathfrak{f}^{-1}\big( \mathfrak{f}(x) \cdot \mathfrak{f}(y) \big),$$
-where $\cdot$ denotes scalar (element-wise) product.
-
-Note that, for the computation of $(\_ \star \_),$ the specific order in which the direct transform
-$\mathfrak{f}$ produces its outputs, and the order in which the inverse transform
-$\mathfrak{f}^{-1}$ expects its inputs to be in, do not matter; we only need these two orders
-to match.
+Note that in~\eqref{eq:conv_theorem}, for the computation of $(\_ \star \_),$ the specific order in
+which the direct transform $\mathfrak{f}$ produces its outputs, and the order in which the inverse
+transform $\mathfrak{f}^{-1}$ expects its inputs to be in, do not matter; we only need these two
+orders to match.
 Indeed, $(\_ \cdot \_)$ is element-wise, $(x \cdot y)[i] = x[i] y[i],$
-so, for any permutation $\pi$ we have $$\pi^{-1}(\pi(x) \cdot \pi(y)) = x \cdot y.$$
+so, for any permutation $\pi$ we have
+\begin{equation} \label{eq:cdot_permutation_invariance}
+\pi^{-1}(\pi(x) \cdot \pi(y)) = x \cdot y.
+\end{equation}
 
@@
 Then, given $j$ and $m=2^s,$ the factor $\varphi$ can be expressed as
-$T_\ast[\Delta_s + j],$ where $\Delta_s = 2^{s-1}-1 = m/2 - 1.$
+$T_\ast[\delta + j],$ where $\delta = 2^{s-1}-1 = m/2 - 1.$
 
@@
 
-We are now returning to the Montgomery reduction.
-
-Remember we defined $R = 2^w \in \mathrm{F}_p,$ where $w$ is the length of machine word in bits;
-we assume $p > 2,$ so $R$ is non-zero in $\mathrm{F}_p.$
-
-Remember also we defined the function
-$\redc \colon \mathrm{F}_p \times \mathrm{F}_p \to \mathrm{F}_p$
-in the following way:
-$$\redc(x, y) = R^{-1} x y.$$
+We are now returning to the Montgomery reduction; we defined $R$ and $\redc$
+in~\ref{section:montgomery_intro}; for the purposes of this section, $p > 2$ is arbitrary prime.
 
@@
 We have the following identities, by the isomorphism, for any $x \in \mathrm{F}_p^N$:
-\begin{eqnarray*}
-    \mathfrak{f}(x)         &=& R^{-1} \, \widetilde{\mathfrak{f}}(R \, x); \\
-    \mathfrak{f}^{-1}(x)    &=& R^{-1} \, \widetilde{\mathfrak{f}}^{-1}(R \, x).
-\end{eqnarray*}
+\begin{subequations}
+\begin{align}
+    \mathfrak{f}(x)         &= R^{-1} \, \widetilde{\mathfrak{f}}(R \, x); \\
+    \mathfrak{f}^{-1}(x)    &= R^{-1} \, \widetilde{\mathfrak{f}}^{-1}(R \, x).
+\end{align}
+\end{subequations}
 
 Rewrite with $x = R^{-1}y$:
-\begin{eqnarray*}
-    \mathfrak{f}(R^{-1}y)       &=& R^{-1} \, \widetilde{\mathfrak{f}}(y); \\
-    \mathfrak{f}^{-1}(R^{-1}y)  &=& R^{-1} \, \widetilde{\mathfrak{f}}^{-1}(y).
-\end{eqnarray*}
-
-Rewrite the left-hand sides by the linearity of $\mathfrak{f}$ and $\mathfrak{f}^{-1}$:
-\begin{eqnarray*}
-    R^{-1} \mathfrak{f}(y)       &=& R^{-1} \, \widetilde{\mathfrak{f}}(y); \\
-    R^{-1} \mathfrak{f}^{-1}(y)  &=& R^{-1} \, \widetilde{\mathfrak{f}}^{-1}(y).
-\end{eqnarray*}
-
-Dividing both sides by $R^{-1},$ we see that $\mathfrak{f}(y)$ and $\widetilde{\mathfrak{f}}(y)$
-coincide for any $y = Rx;$ and so do $\mathfrak{f}^{-1}(y)$ and $\widetilde{\mathfrak{f}}^{-1}(y).$
+\begin{subequations}
+\begin{align}
+    \mathfrak{f}(R^{-1}y)       &= R^{-1} \, \widetilde{\mathfrak{f}}(y); \\
+    \mathfrak{f}^{-1}(R^{-1}y)  &= R^{-1} \, \widetilde{\mathfrak{f}}^{-1}(y).
+\end{align}
+\end{subequations}
+
+Rewrite the left-hand sides by the linearity of $\mathfrak{f}$ and $\mathfrak{f}^{-1}$
+(see~\eqref{eq:transform_linearity}):
+\begin{subequations}
+\begin{align}
+    R^{-1} \mathfrak{f}(y)       &= R^{-1} \, \widetilde{\mathfrak{f}}(y); \\
+    R^{-1} \mathfrak{f}^{-1}(y)  &= R^{-1} \, \widetilde{\mathfrak{f}}^{-1}(y).
+\end{align}
+\end{subequations}
+
+Dividing both sides by $R^{-1},$ we see that the transforms in $\mathrm{F}_p$ and
+$\widetilde{\mathrm{F}}_p$ completely coincide: for any
+$y \in \mathrm{F}_p^N,$
+\begin{subequations} \label{eq:dft_idft_in_tildefp}
+\begin{align}
+\mathfrak{f}(y) &= \widetilde{\mathfrak{f}}(y); \\
+\mathfrak{f}^{-1}(y) &= \widetilde{\mathfrak{f}}^{-1}(y).
+\end{align}
+\end{subequations}
 
 Let us now consider the cyclic convolution $\widetilde{\star}$ for $\widetilde{\mathrm{F}}_p;$
-we know that
+by~\eqref{eq:conv_theorem}, we know that
 $$x \, \widetilde{\star} \, y = \widetilde{\mathfrak{f}}^{-1}
@@
 (which is, remember, the multiplication operation of $\widetilde{\mathrm{F}}_p.$)
-Having proven that
-$\widetilde{\mathfrak{f}} \equiv \mathfrak{f}$
-and
-$\widetilde{\mathfrak{f}}^{-1} \equiv \mathfrak{f}^{-1}$
-above, we can rewrite it as follows:
+Having proven~\eqref{eq:dft_idft_in_tildefp}, we can rewrite it as follows:
 $$x \, \widetilde{\star} \, y = R^{-1} \mathfrak{f}^{-1}\big( \mathfrak{f}(x) \cdot \mathfrak{f}(y) \big),$$
@@
 We now describe the four-step and six-step algorithms for DFT/IDFT, which are matrix Fourier
-algorithms, meaning that they interpret sequences of length $N = N_1 N_2$ as $N_1 \times N_2$ matrices.
+algorithms, meaning that they interpret sequences of length $N = N_1 N_2$ as $N_1 \times N_2$
+matrices.
 
@@
 
-Remember in section~\ref{section:omitting1} we established that, with
-$x \cdot y$ denoting scalar (element-wise) product between $x,y \in \mathcal{F}^N,$
-for any permutation $\pi$ we have $$\pi^{-1}(\pi(x) \cdot \pi(y)) = x \cdot y.$$
-
+Remember in section~\ref{section:omitting1} we established~\eqref{eq:cdot_permutation_invariance}.
 We then used this property to get rid of bit-reversal permutations in DIT and DIF.
@@
 
+Suppose we want to transpose a matrix with $n$ rows and $2n$ columns in-place.
+
+Let $M$ be the matrix we want to transpose.
+Split $M$ into two $n$-by-$n$ submatrices, left submatrix $A$ and right submatrix $B$:
+$$
+M = \begin{bmatrix}
+A & B
+\end{bmatrix}.
+$$
+
+Note that
+$$
+M^T =
+\begin{bmatrix}
+    A & B
+\end{bmatrix}^T =
+\begin{bmatrix}
+    A^T \\
+    B^T
+\end{bmatrix}.
+$$
+
+Let us refer to the rows of $A^T$ as $\alpha_1, \ldots, \alpha_n;$
+and to the rows of $B^T$ as $\beta_1, \ldots, \beta_n.$
+We write $\longleftrightarrow$ for ``equivalent in flat row-major order to''.
+Then,
+$$
+\begin{bmatrix}
+    A^T & B^T
+\end{bmatrix} =
+\begin{bmatrix}
+    \alpha_1 & \beta_1 \\
+    \ldots & \ldots \\
+    \alpha_n & \beta_n
+\end{bmatrix}
+\longleftrightarrow
+\langle \alpha_1, \beta_1, \alpha_2, \beta_2, \ldots, \alpha_n, \beta_n \rangle.
+$$
+It differs from
+$$
+M^T =
+\begin{bmatrix}
+    A^T \\
+    B^T
+\end{bmatrix} =
+\begin{bmatrix}
+    \alpha_1  \\
+    \ldots \\
+    \alpha_n \\
+    \beta_1 \\
+    \ldots \\
+    \beta_n
+\end{bmatrix}
+\longleftrightarrow
+\langle \alpha_1, \alpha_2, \ldots, \alpha_n, \beta_1, \beta_2, \ldots, \beta_n \rangle.
+$$
+
+So we can propose the following algorithm for transposing $M,$ although quite inefficient:
+\begin{enumerate}
+    \item Interpreting the whole array as $n \times 2n$ matrix, transpose the left and right submatrices.
+    \item Interpreting the whole array as $2n \times n$ matrix, permute its \textit{rows} in a certain way.
+Namely, we need to apply the permutation $\rho$ that sends a row with zero-based index
+$i, 0 \le i < 2n,$ to the zero-based index $$n \cdot (i \mymod 2) + (i \mydiv 2).$$
+\end{enumerate}
+
+But permuting \textit{rows} of a matrix means permuting \textit{columns} of the transposed matrix
+in the same way. To write it in a formal way, we need to introduce some new notation.
+
+If $\pi$ is a permutation on length-$N$ sequences,
+then $\pi' X,$ where $X$ is a matrix with $N$ rows, means $X$ with rows permuted by $\pi;$
+$\pi'' Y,$ where $Y$ is a matrix with $N$ columns, means $Y$ with columns permuted by $\pi.$
+
+We can then write formally, for any matrix $X$ with $N$ columns and permutation $\pi$ on length-$N$
+sequences,
+\begin{equation} \label{eq:matrix_permutation_rule}
+(\pi'' X)^T = \pi'(X^T).
+\end{equation}
+
+Define the result of transposition of both left and right submatrices of a $n$-by-$2n$ matrix $M$ as
+$\Psi(M).$ Note this is step 1 of our algorithm. We have:
+$$\rho' \Psi(M) = M^T.$$
+Substitute with $M = \rho'' L:$
+$$\rho' \Psi(\rho'' L) = (\rho'' L)^T.$$
+Apply~\eqref{eq:matrix_permutation_rule} to the right-hand side:
+$$\rho' \Psi(\rho'' L) = \rho'(L^T).$$
+Simplify, since $\rho'$ is a bijection:
+$$\Psi(\rho'' L) = L^T.$$
+
+It means that we can apply $\rho$ to \textit{columns} (as opposed to rows) \textit{before} the transpositions of
+submatrices (as opposed to after) --- and still get the transposed matrix as the result.
+
+Application of $\rho$ to a column can be done in a cache-efficient way, and requires $2n$ additional memory.
+
+Since $(M^T)^T = M,$ the transposition of $2n \times n$ matrix is the inverse operation on the array of
+the same length; it can be done as doing the inverse operations in reverse order: apply $\rho^{-1}$
+to columns, then transpose the submatrices. Application of $\rho^{-1}$ to a column can also be done in a
+cache-efficient way, and also requires $2n$ additional memory.
+
+This approach generalizes trivially to the case of $n \times C n$ and $C n \times n$ matrices.
+It requires $C n$ additional memory, which, for fixed $C,$ is $O(\sqrt{N}),$ where
+$N = C n^2$ is the number of elements in the matrix.
+
 \subsection{Optimizing the four-step algorithm for $N=3 \cdot 2^k$}
@@
 
-\subsection{Power table}
+We want to obtain an upper bound, in machine words, for the value of $\sigma_i + c_i$
+in~\eqref{eq:sigmai_and_zi}.
+Define $L = (B-1)^2 M.$ By~\eqref{eq:product_inequality} and~\eqref{eq:pi_bound}, we have
+$$L < p_1 p_2 < (\mu / 2)^2 = \mu^2 / 4.$$
+
+By~\eqref{eq:ci_bound}, we have $c_i \le L;$ and by~\eqref{eq:sigmai_bound}, we have
+$\sigma_i < (B-1)M \le L.$ But then
+$$\sigma_i + c_i < 2L < \mu^2 / 2.$$
+
+Note this is far from being a tight bound in numerical sense:
+in practice, the value of
+$\frac{L}{(B-1)M}=B-1$ is going to be much larger than 1, meaning that the exact upper bound on
+$c_i + \sigma_i$ is going to be closer to $\mu^2 / 4$ than to $\mu^2 / 2;$ but this is still a
+tight bound in the sense that maximum possible $c_i + \sigma_i$ would not into one machine word.
+This bound also has the advantage that it works for any $B > 1.$
 

### 2020-11-18 fft.c

@@ -432,8 +432,9 @@ void convolve_fourstep(
         FFT_ULIMB *scratch,
         const Field *field)
 {
-    size_t c = n / 3;
-    unsigned order = fft_zu_counttz(c);
+    // Odd number times 2^k has exactly k trailing zeros in binary.
+    unsigned order = fft_zu_counttz(n);
+    size_t c = ((size_t) 1) << order;
 
     // make table of powers for root (in Montgomery representation)
     mk_special_pow_table(field->root_M, scratch, c, field);

### 2020-11-18 paper/paper.tex
Update

@@
 bit-reversal permutations for step 5 of the direct transform and DIT without bit-reversal
-permutations for step 2 of the inverse transform; but we still have two matrix transposition steps
+permutations for step 2 of the inverse transform. But we still have two matrix transposition steps
 per transform, and no matter whether we use DIT or DIF for step 2 of the direct transform
@@
 
+\subsection{Overview of matrix transposition}
+
+We will now discuss matrix transposition.
+Matrix transposition can be done either out-of-place or in-place.
+
+Out-of-place transposition means a separate array in the desired order is produced; this means
+using $O(N_1 N_2)$ additional memory, where $N_1$ and $N_2$ are dimensions of the matrix.
+
+The definitions of in-place transposition vary, but generally it is defined as an algorithm that
+uses ``much less'' additional memory than $O(N_1 N_2).$
+
+Another property that we might want from a transposition algorithm is cache friendliness.
+
+Note that a cache-oblivious algorithm for out-of-place matrix transposition is
+well-known~\cite{cache_rect_mat_trans}.
+We now want to explore the space of in-place transposition algorithms.
+
+The algorithm for in-place transposition of a square matrix is simple: element
+$\langle i, j \rangle$ is exchanged with element $\langle j, i \rangle.$
+The same approach works for transposing a square submatrix of a larger, generally rectangular
+matrix.
+For completeness, we provide the code below.
+
+\begin{procedure}[H]
+\caption{TransposeSquareSubMatrix($p$, $n$, $n'$)}
+\DontPrintSemicolon
+\KwData{
+    Pointer $p$ to the first row, first column of $n$-by-$n$ submatrix of a possibly larger matrix
+    with $n''$ columns and $n'$ rows stored in row-major order.
+    (Formally, if the submatrix starts at $i$-th row, $0 \le i < n'$ and
+    $j$-th column, $0 \le j < n'',$ then $p = q + n' \cdot i + j,$ where $q$ is the pointer to the
+    beginning of the larger matrix.)
+
+    Integer $n.$
+
+    Integer $n'.$
+}
+\KwResult{
+    The square submatrix is transposed.
+}
+\Begin{
+    \For{$i = 0$ \KwTo $n-1$}{
+        \For{$j = i + 1$ \KwTo $n-1$}{
+            {exchange $p[i \cdot n' + j]$ with $p[j \cdot n' + i]$}\;
+        }
+    }
+}
+\end{procedure}
+
+For a cache-oblivious version of this algorithm, see~\cite{cache_square_mat_trans}.
+
+We will now discuss in-place transposition of a rectangular matrix.
+
+In its general form, in-place transposition of a non-square matrix is quite hard.
+The classical study begins with defining the function $\mathcal{P}(a)$ such that,
+when transposing a matrix with $N_1$ rows and $N_2$ columns in row-major order,
+the element with index $a$ is sent to index $\mathcal{P}(a).$
+There is, then, a result which says that
+\begin{equation}
+\mathcal{P}(a) = \begin{cases}
+    N_1 N_2 - 1                 & \text{if } a = N_1 N_2 - 1; \\
+    N_1 a \mymod (N_1 N_2 - 1)  & \text{otherwise.}
+\end{cases}
+\end{equation}
+The number of fixed points of $\mathcal{P}$ is exactly
+$$
+1 + \gcd(N_1-1, N_1-1).
+$$
+The number of cycles of length $k>1$ of $\mathcal{P}$ is given by
+$$
+\frac{1}{k} \sum\limits_{d | k} \widetilde{\mu}(k/d) \gcd(N_1^d - 1, N_1 N_2 - 1),
+$$
+where $\widetilde{\mu}$ is the Möbius function and the summation is performed over all divisors of
+$k.$
+
+Most algorithms for in-place transposition are essentially of ``follow-the-cycles'' kind.
+This means that they iterate over all the cycles, and for each cycle $C,$
+they shift the elements of $C$ cyclically. There are, then, various approaches for
+locating the cycles, identifying the first element of a cycle, and tracking which cycles were
+already visited.
+If we limit ourselves to using only $O(N_1+N_2)$ additional memory,
+even more complex algorithms are genreally needed;
+\cite{Fich95permutingin}
+presents an algorithm with worst-case complexity of $O(N_1 N_2 \log(N_1 N_2)).$
+
+Notably, the algorithm for matrix transposition in \textbf{mpdecimal} is also
+``follow-the-cycles'' in nature, requiring $O(N_1 N_2)$ additional memory, although with a
+small constant: it uses a bit set to track which elements were already shifted.
+
+In the circumstances described above, one would believe,
+we have no choice but to give up and accept the complexity --- after all,
+much research has been dedicated to this problem;
+if there were a simple solution, somebody would find it --- and perhaps
+\textbf{mpdecimal} would use it.
+After all, the case of six-step FFT with its
+$2^k \times 2^{k+1}$ or $2^{k+1} \times 2^k$ matrices must be an important enough
+application for a matrix transposition.
+
+Well, as it turns out, there \textit{is} a simple solution,
+even more general than we need: it works for
+$n \times Cn$ or $Cn \times n$ matrices, for any constant $C.$
+It requires $Cn = O(n) = O(N_1) = O(N_2)$ additional memory,
+has $O(N_1 N_2)$ time complexity, and is very cache-friendly,
+assuming a cache-friendly procedure for transposing a square matrix is available.
+
 \subsection{Algorithm for matrix transposition}
 
+We will now describe our algorithm for in-place transposition of $n \times Cn$ matrix,
+first for $C=2.$
+We will then show how to perform in-place transposition of $2n \times n$ matrix,
+which is the inverse operation.
+Finally, we discuss how our approach can be generalized for arbitrary $C.$
+
 Suppose we want to transpose a matrix with $n$ rows and $2n$ columns in-place.
-
 Let $M$ be the matrix we want to transpose.
-Split $M$ into two $n$-by-$n$ submatrices, left submatrix $A$ and right submatrix $B$:
+
+We split $M$ into two $n$-by-$n$ submatrices, and name the left submatrix $A$ and the right
+submatrix $B$:
 $$
@@
 
+Define $\Psi(M)$ to be the result of transposing the left and right submatrices of $M$:
+$$
+\Psi(M) = \begin{bmatrix}
+    A^T & B^T
+\end{bmatrix}.
+$$
+Note $\Psi(M) \ne M^T;$
+we now want to explore how exactly $\Psi(M)$ differs from $M^T$ when both are written as ``flat''
+sequences in row-major order.
+
 Let us refer to the rows of $A^T$ as $\alpha_1, \ldots, \alpha_n;$
 and to the rows of $B^T$ as $\beta_1, \ldots, \beta_n.$
-We write $\longleftrightarrow$ for ``equivalent in flat row-major order to''.
+We write $\longleftrightarrow$ for ``equivalent in row-major order flat form to''.
 Then,
 $$
-\begin{bmatrix}
+\Psi(M) = \begin{bmatrix}
     A^T & B^T
@@
 
-So we can propose the following algorithm for transposing $M,$ although quite inefficient:
+So we can propose the following algorithm for transposing $M,$ although not yet in-place:
 \begin{enumerate}
@@
 then $\pi' X,$ where $X$ is a matrix with $N$ rows, means $X$ with rows permuted by $\pi;$
-$\pi'' Y,$ where $Y$ is a matrix with $N$ columns, means $Y$ with columns permuted by $\pi.$
+and $\pi'' Y,$ where $Y$ is a matrix with $N$ columns, means $Y$ with columns permuted by $\pi.$
 
@@
 
-Define the result of transposition of both left and right submatrices of a $n$-by-$2n$ matrix $M$ as
-$\Psi(M).$ Note this is step 1 of our algorithm. We have:
+We have:
 $$\rho' \Psi(M) = M^T.$$
-Substitute with $M = \rho'' L:$
+Substitute with $M = \rho'' L$:
 $$\rho' \Psi(\rho'' L) = (\rho'' L)^T.$$
@@
 
-It means that we can apply $\rho$ to \textit{columns} (as opposed to rows) \textit{before} the transpositions of
-submatrices (as opposed to after) --- and still get the transposed matrix as the result.
+It means that we can apply $\rho$ to \textit{columns} (as opposed to rows) \textit{before} the
+transpositions of submatrices (as opposed to after) --- and still get the transposed matrix as the
+result.
 
@@
 the same length; it can be done as doing the inverse operations in reverse order: apply $\rho^{-1}$
-to columns, then transpose the submatrices. Application of $\rho^{-1}$ to a column can also be done in a
+to columns, then transpose the submatrices. Although $\rho^{-1} \ne \rho,$
+application of $\rho^{-1}$ to a column can also be done in a
 cache-efficient way, and also requires $2n$ additional memory.
 
-This approach generalizes trivially to the case of $n \times C n$ and $C n \times n$ matrices.
-It requires $C n$ additional memory, which, for fixed $C,$ is $O(\sqrt{N}),$ where
-$N = C n^2$ is the number of elements in the matrix.
+We provide the pseudocode for case of $C=2$ below.
+
+Note it uses the routine $\mathrm{TransposeSquareSubMatrix}$.
+It does not have to be implemented exactly in a way we demonstrated above;
+it just needs to have the same semantics.
+Note that it can be implemented in a cache-friendly way by dividing the square submatrix into
+tiles, perhaps recursively, transposing the elements inside tiles, and then transposing the tiles
+themselves; see~\cite{cache_square_mat_trans} for details.
+
+\begin{procedure}[H]
+\caption{PermuteRho($p$, $\zeta$, $n$)}
+\DontPrintSemicolon
+\KwData{
+    Pointer $p$ to array of size $2n$.
+    Pointer $\zeta$ to scratch buffer of size $2n.$
+    Integer $n.$
+}
+\KwResult{
+    The permutation $\rho$ is applied to the array.
+    The contents of the scratch buffer is undefined.
+}
+\Begin{
+    \For{$i = 0$ \KwTo $2n-1$}{
+        $\zeta[i] \longleftarrow p[i]$\;
+    }
+    \For{$i = 0$ \KwTo $n-1$}{
+        $p[i] \longleftarrow \zeta[2i]$\;
+        $p[i+n] \longleftarrow \zeta[2i+1]$\;
+    }
+}
+\end{procedure}
+
+\begin{procedure}[H]
+\caption{UnPermuteRho($p$, $\zeta$, $n$)}
+\DontPrintSemicolon
+\KwData{
+    Pointer $p$ to array of size $2n$.
+    Pointer $\zeta$ to scratch buffer of size $2n.$
+    Integer $n.$
+}
+\KwResult{
+    The permutation $\rho^{-1}$ is applied to the array.
+    The contents of the scratch buffer is undefined.
+}
+\Begin{
+    \For{$i = 0$ \KwTo $2n-1$}{
+        $\zeta[i] \longleftarrow p[i]$\;
+    }
+    \For{$i = 0$ \KwTo $n-1$}{
+        $p[2i] \longleftarrow \zeta[i]$\;
+        $p[2i+1] \longleftarrow \zeta[i+n]$\;
+    }
+}
+\end{procedure}
+
+\begin{procedure}[H]
+\caption{TransposeMatrixNx2N($p$, $\zeta$, $n$)}
+\DontPrintSemicolon
+\KwData{
+    Pointer $p$ to a matrix with $n$ rows and $2n$ columns stored in row-major order.
+    Pointer $\zeta$ to scratch buffer of size $2n.$
+    Integer $n.$
+}
+\KwResult{
+    The matrix is transposed.
+    The contents of the scratch buffer is undefined.
+}
+\Begin{
+    \For{$i = 0$ \KwTo $n-1$}{
+        $\mathrm{PermuteRho}(p + i \cdot 2n, \zeta, n)$\;
+    }
+    $\mathrm{TransposeSquareSubMatrix}(p, n, 2n)$\;
+    $\mathrm{TransposeSquareSubMatrix}(p + n, n, 2n)$\;
+}
+\end{procedure}
+
+\begin{procedure}[H]
+\caption{TransposeMatrix2NxN($p$, $\zeta$, $n$)}
+\DontPrintSemicolon
+\KwData{
+    Pointer $p$ to a matrix with $2n$ rows and $n$ columns stored in row-major order.
+    Pointer $\zeta$ to scratch buffer of size $2n.$
+    Integer $n.$
+}
+\KwResult{
+    The matrix is transposed.
+    The contents of the scratch buffer is undefined.
+}
+\Begin{
+    $\mathrm{TransposeSquareSubMatrix}(p, n, 2n)$\;
+    $\mathrm{TransposeSquareSubMatrix}(p + n, n, 2n)$\;
+    \For{$i = 0$ \KwTo $n-1$}{
+        $\mathrm{UnPermuteRho}(p + i \cdot 2n, \zeta, n)$\;
+    }
+}
+\end{procedure}
+
+This approach generalizes trivially to the case of $n \times C n$ and $C n \times n$ matrices;
+it would require $Cn$ additional memory.
 

### 2020-11-19 paper/paper.tex
Update

@@
     (1) the ``linearity trick'' for using Montgomery reduction with NTT without separate steps of conversion into/out of Montgomery form;
-    (2) a simple cache-oblivious algorithm for in-place $2n \times n$ or $n \times 2n$ matrix transposition, the need for which arises in the ``six-step algorithm'' variation of the matrix Fourier algorithm, and which does not seem to be widely known.
+    (2) a simple cache-efficient algorithm for in-place $2n \times n$ or $n \times 2n$ matrix transposition, the need for which arises in the ``six-step algorithm'' variation of the matrix Fourier algorithm, and which does not seem to be widely known.
     Notable findings include:
@@
 
-Unfortunately, this method is very slow. % TODO
+Unfortunately, this method is very slow; in our benchmark, native division on x86-64
+was 720\% slower than the Montgomery reduction. The code can be found in the
+\verb!bench-modmul! subdirectory of our repository.
 
@@
 $\approx$ 15\% compared to Montgomery reduction;
-the code can be found in the \verb!bench-solinas! subdirectory of our repository.
+the code can be found in the \verb!bench-modmul! subdirectory of our repository.
 
@@
 
-IDFT of length $N = RC$ can be computed as follows:
+IDFT of length $N = RC$ can be computed as follows, not counting final multiplication by $N^{-1}$:
 \begin{enumerate}
@@
 $$
-1 + \gcd(N_1-1, N_1-1).
+1 + \gcd(N_1-1, N_2-1).
 $$
@@
 Most algorithms for in-place transposition are essentially of ``follow-the-cycles'' kind.
-This means that they iterate over all the cycles, and for each cycle $C,$
-they shift the elements of $C$ cyclically. There are, then, various approaches for
+This means that they iterate over all the cycles, and for each cycle,
+they shift its elements cyclically. There are, then, various approaches for
 locating the cycles, identifying the first element of a cycle, and tracking which cycles were
@@
 \subsection{Algorithm for matrix transposition}
+\label{section:mat_trans_algo}
 
@@
 
-Application of $\rho$ to a column can be done in a cache-efficient way, and requires $2n$ additional memory.
+Application of $\rho$ to a column can be done in a cache-efficient way, and requires $2n$
+additional memory.
 
@@
 it just needs to have the same semantics.
-Note that it can be implemented in a cache-friendly way by dividing the square submatrix into
-tiles, perhaps recursively, transposing the elements inside tiles, and then transposing the tiles
+It can be implemented in a cache-friendly way by dividing the square submatrix into
+tiles, perhaps recursively, transposing the elements inside tiles, and then swapping the tiles
 themselves; see~\cite{cache_square_mat_trans} for details.
@@
 
+With $N=3 \cdot 2^k,$ we interpret the sequence as a matrix with 3 rows and $2^k$ columns.
+
+Observe that steps 1 and 2 of the direct transform, namely
+\begin{itemize}
+\item length-3 DFT on each column,
+\item multiplication by $(\xi_N)^{ij}$),
+\end{itemize}
+as well as steps 3 and 4 of the inverse transform, namely
+\begin{itemize}
+\item multiplication by $(\xi_N)^{-ij},$
+\item length-3 IDFT on each column,
+\end{itemize}
+can be merged (performed in one pass over the columns).
+In fact, they should be merged, as it produces less loads/stores and leads to better
+cache utilization.
+
+If, additionally, we need to multiply each matrix element by some factor $\Phi,$
+then we can meld this operation into the column operations as well; and, as it turns out,
+this can save us a few multiplications per column.
+Note there is no difference at which step to multiply the matrix by a scalar $\Phi,$
+because all the column operations involved are linear.
+
+Note that we do need to multiply everything by $N^{-1}$ after performing the inverse transform
+(or by another factor if the linearity trick is used; see section~\ref{section:trick}).
+For the direct transform, we only need to ``multiply'' (perform $\redc$ with $R^2$)
+if the linearity trick is not used.
+
+Suppose we need to perform length-3 DFT on each column, then multiply each matrix element indexed
+$\langle i,j \rangle$
+by $(\xi_N)^{ij};$ and also multiply everything by a factor $\Phi.$
+Define $\Gamma = \xi_N,$ $\Lambda = (\xi_N)^{N/3}.$
+This operation on a signle column with index $j,$
+$\langle x, y, z \rangle \mapsto \langle x', y', z' \rangle,$ can be expresed as:
+
+\begin{subequations}
+\begin{align}
+    x' &= \Phi (x + y + z); \\
+    y' &= \widetilde{\Phi}_1 (x + y \Lambda   + z \Lambda^2); \\
+    z' &= \widetilde{\Phi}_2 (x + y \Lambda^2 + z \Lambda),
+\end{align}
+\end{subequations}
+where
+\begin{subequations}
+\begin{align}
+    \widetilde{\Phi}_1 &= \Phi \Gamma^j; \\
+    \widetilde{\Phi}_2 &= \Phi \Gamma^{2j}.
+\end{align}
+\end{subequations}
+
+Note that $\widetilde{\Phi}_1$ and $\widetilde{\Phi}_2$ can be computed incrementally,
+with two variables initially having the value of $\Phi$ and get multiplied by
+$\Gamma$ and $\Gamma^2,$ correspondingly, after each iteration of the loop as we operate on columns
+$0, \ldots, 2^k-1.$ This is an improvement over a separate multiply-by-$\Phi$ step:
+we have 9 multiplications per column instead of 11.
+Note that if we do not need to multiply by $\Phi,$ we have 8 multiplications per column.
+
+Suppose now we need to multiply each matrix element indexed $\langle i,j \rangle$
+by $(\xi_N)^{-ij},$ then perform length-3 IDFT on each column; and also multiply everything by a
+factor $\Phi.$
+
+Define $\Gamma = (\xi_N)^{-1},$ $\Lambda = (\xi_N)^{-N/3}.$
+This operation on a signle column with index $j,$
+$\langle x, y, z \rangle \mapsto \langle x', y', z' \rangle,$ can be expresed as:
+
+\begin{subequations}
+\begin{align}
+    x' &= \widetilde{x} + \widetilde{y}           + \widetilde{z}; \\
+    y' &= \widetilde{x} + \widetilde{y} \Lambda   + \widetilde{z} \Lambda^2; \\
+    z' &= \widetilde{x} + \widetilde{y} \Lambda^2 + \widetilde{z} \Lambda,
+\end{align}
+\end{subequations}
+where
+\begin{subequations}
+\begin{align}
+    \widetilde{x} &= x \cdot \Phi; \\
+    \widetilde{y} &= y \cdot \Phi \Gamma^j; \\
+    \widetilde{z} &= z \cdot \Phi \Gamma^{2j}.
+\end{align}
+\end{subequations}
+
+Again, the factors for $\widetilde{y}$ and $\widetilde{z},$
+that is $\Phi \Gamma^j$ and $\Phi \Gamma^{2j},$ can be computed incrementally.
+Similarly to the case of direct transform, this is an improvement over a
+separate multiply-by-$\Phi$ step: 9 multiplications per column instead of 11.
+
 \subsection{Optimizing the six-step algorithm for $N=2^k$}
+
+We need to factorize $N = 2^k$ into $N_1 \cdot N_2 = N$ so that $\max\{N_1, N_2\}$ is as small as
+possible.
+This is because we will perform transforms of lengths both $N_1$ and $N_2,$ and we want to achieve
+good cache utilization.
+This is done by choosing $N_1 = 2^{\lfloor k / 2 \rfloor}, N_2 = 2^{\lceil k / 2 \rceil}.$
+Note that in case of even $k,$ the matrix is square; in case of odd $k,$
+$N_2 = 2 \cdot N_1$ and the matrix can be transposed with the algorithm we described in
+section~\ref{section:mat_trans_algo}.
+
+Just like with the four-step algorithm, it is desirable to merge together the steps of
+multiplication by $(\xi_N)^{\pm ij}$ and multiplication by a scalar $\Phi,$
+and compute the factor $\Phi (\xi_N)^{\pm ij}$ for the current element at
+$\langle i,j \rangle$ incrementally.
+Compared to a separate multiply-by-$\Phi$ step, this approach
+lowers the number of multiplications from $3N + o(N)$ to $2N + o(N).$
 
@@
 $c_i + \sigma_i$ is going to be closer to $\mu^2 / 4$ than to $\mu^2 / 2;$ but this is still a
-tight bound in the sense that maximum possible $c_i + \sigma_i$ would not into one machine word.
+tight bound in the sense that the maximum possible $c_i + \sigma_i$ would overflow one machine word.
 This bound also has the advantage that it works for any $B > 1.$
@@
 Remember the procedure $\mathrm{MontgomeryReduce}(a)$ that we defined in section~\ref{section:montgomery_intro}.
+
+\subsection{Cache aliasing problems}
 

### 2020-11-19 bench-modmul/bench.c

@@ -14,7 +14,34 @@ DOUBLE_ULIMB full_prod(ULIMB x, ULIMB y)
     return ((DOUBLE_ULIMB) x) * y;
 }
 
-#if defined(USE_SOLINAS)
+#if defined(USE_NAIVE)
+
+#define MODULO 18446744069414584321ull
+static inline FORCE_INLINE
+ULIMB mul(ULIMB x, ULIMB y)
+{
+    DOUBLE_ULIMB t = full_prod(x, y);
+    ULIMB t_hi = t >> LIMB_BITS;
+    ULIMB t_lo = t;
+
+    ULIMB q, r;
+
+    asm inline (
+        "divq %4\n"
+
+        : /*0*/ "=a" (q)
+        , /*1*/ "=d" (r)
+
+        : /*2*/ "0" (t_lo)
+        , /*3*/ "1" (t_hi)
+        , /*4*/ "r" (MODULO)
+
+        : "cc");
+
+    return r;
+}
+
+#elif defined(USE_SOLINAS)
 
 #define MODULO 18446744069414584321ull
 static inline FORCE_INLINE
@@ -55,7 +82,7 @@ ULIMB mul(ULIMB x, ULIMB y)
 }
 
 #else
-#   error "Please define either USE_SOLINAS or USE_MONT."
+#   error "Please define either of: USE_NAIVE, USE_SOLINAS, USE_MONT."
 #endif
 
 static int parse_int_or_die(const char *s)

### 2020-11-21 paper/paper.tex
Update

@@
 \usepackage{cite}
+\usepackage{url}
 \usepackage[ruled]{algorithm2e}
@@
 a group of algorithms with different asymptotic behavior such that it makes sense to use each one of them only for some range of values of $n.$
-To give an example, \textbf{libgmp} employs the following ``hierarchy of fanciness'' for multiplication:
+To give an example, \textbf{libgmp} employs the following ``hierarchy of fanciness'' for multiplication~\cite{gmp_fh}:
 first goes the ``basecase'' (quadratic) algorithm; then, the Karatsuba algorithm; then, variations of the Toom-Cook algorithm (Toom-3; then Toom-4; then Toom-6.5; then Toom-8.5); finally, the Schönhage--Strassen algorithm (which is Fourier transform-based).
@@
 
-The convolution theorem says that, for any two vectors $x, y \in \mathcal{F}^N,$
+The convolution theorem~\cite{conv_theorem} says that, for any two vectors $x, y \in \mathcal{F}^N,$
 \begin{equation} \label{eq:conv_theorem}
@@
 
-Although an $O(N \log N)$ FFT algorithm exists that works for arbitrary $N,$ even prime,
+Although an $O(N \log N)$ FFT algorithm exists that works for arbitrary $N,$ even
+prime~\cite{bluestein},
 the most widely used FFT algorithm, the Cooley---Tukey algorithm, works by re-writing a transform
@@
 some point, you are going to get a round-off error large enough to produce a wrong digit in the
-answer.
+answer~\cite{precision_paper}.
 
 In order to get a provably correct answer, you have to put less bits of information into the
-floating-point values than otherwise possible. Because of that, in order to achieve performance
+floating-point values than otherwise possible~\cite{gmp_paper, mattcomp_precision}.
+Because of that, in order to achieve performance
 parity with multiplication algorithms that abstain from use of floating point, platform-specific
-SIMD extensions are often used. Also, in order to get a provably correct result, you need to know
-the precision of your floating-point types, as well as possible quirks of their implementation ---
-i.e., depend on the platform.
+SIMD extensions are often used~\cite{gmp_paper}.
+Also, in order to get a provably correct result, you need to know the precision of your
+floating-point types, as well as possible quirks of their implementation --- i.e., depend on the
+platform.
 
@@
 \subsection{Choosing the base}
+\label{section:bases}
 
@@
 The naïve method of modular reduction is to use, where it is available, a hardware instruction
-to divide two-words divident by single-word divisor into single-word quotient and single-word
+to divide two-words dividend by single-word divisor into single-word quotient and single-word
 remainder. Where not available, we would have to emulate such an instruction in software.
@@
 A method intended to be faster, while not requiring a change in our representation of field
-elements, is known as Barrett reduction. % TODO cite
+elements, is known as Barrett reduction~\cite{barrett}.
 
@@
 
-A method even faster for our purposes, is Montgomery reduction.
+A method even faster for our purposes, is Montgomery reduction~\cite{Montgomery1985ModularMW}.
 
@@
 it is based on the use of prime moduli of form $2^{64} - 2^{n} + 1.$
-Generally, such primes are knows as Solinas primes, defined as primes of form
-$f(2^m),$ where $f(x)$ is a low-degree polynomial with small integer coefficients.
+Generally, such primes are knows as Solinas primes, or generalized Mersenne
+primes~\cite{Solinas},
+defined as primes of form $f(2^m),$ where $f(x)$ is a low-degree polynomial with small integer
+coefficients.
 
@@
 
-DFT of length $N = RC$ can be computed as follows:
+DFT of length $N = RC$ can be computed as follows~\cite{mattcomp_mfa}:
 \begin{enumerate}
@@
 The six-step algorithm has the advantage that, aside from matrix transpositions, it only accesses
-memory in strides of $R$ or $C.$ Thus, it might be faster in settings of hierarchical memory,
+memory in strides of $R$ and $C.$ Thus, it might be faster in settings of hierarchical memory,
 including external-memory transforms and modern systems with multiple levels of cache.
@@
 the element with index $a$ is sent to index $\mathcal{P}(a).$
-There is, then, a result which says that
+It can be defined explicitly as follows:
 \begin{equation}
@@
 \end{equation}
-The number of fixed points of $\mathcal{P}$ is exactly
-$$
-1 + \gcd(N_1-1, N_2-1).
-$$
-The number of cycles of length $k>1$ of $\mathcal{P}$ is given by
+There is, then, a result~\cite{fancything} saying that
+the number of fixed points of $\mathcal{P}$ is exactly
+$$
+1 + \gcd(N_1-1, N_2-1);
+$$
+and the number of cycles of length $k>1$ of $\mathcal{P}$ is
 $$
@@
 If we limit ourselves to using only $O(N_1+N_2)$ additional memory,
-even more complex algorithms are genreally needed;
+even more complex algorithms are generally needed;
 \cite{Fich95permutingin}
@@
 
-Observe that steps 1 and 2 of the direct transform, namely
+Observe that steps 1 and 2 of the direct transform:
 \begin{itemize}
 \item length-3 DFT on each column,
-\item multiplication by $(\xi_N)^{ij}$),
+\item multiplication by $(\xi_N)^{ij},$
 \end{itemize}
-as well as steps 3 and 4 of the inverse transform, namely
+as well as steps 3 and 4 of the inverse transform:
 \begin{itemize}
@@
 \end{itemize}
-can be merged (performed in one pass over the columns).
-In fact, they should be merged, as it produces less loads/stores and leads to better
+can be merged --- performed in a single pass over the columns.
+In fact, they \textit{should} be merged, as it produces less loads/stores and leads to better
 cache utilization.
@@
 
-Note that we do need to multiply everything by $N^{-1}$ after performing the inverse transform
-(or by another factor if the linearity trick is used; see section~\ref{section:trick}).
+Note that we are going to need to multiply everything by $N^{-1}$ after performing the inverse
+transform (or by another factor if the linearity trick is used; see section~\ref{section:trick}).
 For the direct transform, we only need to ``multiply'' (perform $\redc$ with $R^2$)
@@
 
+First, consider the case of direct transform.
 Suppose we need to perform length-3 DFT on each column, then multiply each matrix element indexed
@@
 by $(\xi_N)^{ij};$ and also multiply everything by a factor $\Phi.$
+
 Define $\Gamma = \xi_N,$ $\Lambda = (\xi_N)^{N/3}.$
-This operation on a signle column with index $j,$
-$\langle x, y, z \rangle \mapsto \langle x', y', z' \rangle,$ can be expresed as:
-
+This operation on a single column with index $j,$
+$\langle x, y, z \rangle \mapsto \langle x', y', z' \rangle,$ can be expressed as:
 \begin{subequations}
@@
 
-Suppose now we need to multiply each matrix element indexed $\langle i,j \rangle$
+Now, the case of inverse transform.
+We now need to multiply each matrix element indexed $\langle i,j \rangle$
 by $(\xi_N)^{-ij},$ then perform length-3 IDFT on each column; and also multiply everything by a
@@
 Define $\Gamma = (\xi_N)^{-1},$ $\Lambda = (\xi_N)^{-N/3}.$
-This operation on a signle column with index $j,$
-$\langle x, y, z \rangle \mapsto \langle x', y', z' \rangle,$ can be expresed as:
-
+This operation on a single column with index $j,$
+$\langle x, y, z \rangle \mapsto \langle x', y', z' \rangle,$ can be expressed as:
 \begin{subequations}
@@
 
-We need to factorize $N = 2^k$ into $N_1 \cdot N_2 = N$ so that $\max\{N_1, N_2\}$ is as small as
-possible.
-This is because we will perform transforms of lengths both $N_1$ and $N_2,$ and we want to achieve
-good cache utilization.
-This is done by choosing $N_1 = 2^{\lfloor k / 2 \rfloor}, N_2 = 2^{\lceil k / 2 \rceil}.$
+Remember that the reason we might turn to using the six-step algorithm is its cache
+friendliness: if we interpret the sequence as $N_1 \times N_2$ matrix, then, aside from matrix
+transpositions, it only accesses memory in strides of $N_1$ and $N_2.$
+
+With $N = 2^k,$
+we need to factorize $N$ into $N_1 N_2 = N$ so that $N_1$ and $N_2$ are as close to each other
+as possible.
+
+This can be done by choosing $N_1 = 2^{\lfloor k / 2 \rfloor}, N_2 = 2^{\lceil k / 2 \rceil}.$
 Note that in case of even $k,$ the matrix is square; in case of odd $k,$
@@
 
-\subsection{Modular addition and subtraction}
+\subsection{Modular addition, subtraction, and adjustment}
+
+For the purposes of this subsection, ``adjustment'' means reducing a value in range $[0; 2p-1]$
+modulo $p$ to be in range $[0; p-1].$
+
+Using actual conditional jumps on assembly level for modular addition, subtraction, or adjustment,
+is slow on most modern architectures:
+modern processors employ branch prediction, and branches of this
+sort are anything but predictable. This results in lots of costly branch mispredictions.
+
+We describe two possible ways to resolve this problem.
+
+\subsubsection{Bit wizardry}
+
+We assume here that two's complement is used,
+all values are $w$-bit,
+and a modulo $p < 2^w / 2$ is fixed.
+
+For a signed $w$-bit value $x,$ we define $\underline{\mathrm{ExtractSign}}(x)$ as follows:
+$$
+\underline{\mathrm{ExtractSign}}(x) =
+\begin{cases}
+0, & \text{if } x \ge 0; \\
+-1, & \text{if } x < 0.
+\end{cases}
+$$
+We can compute $\underline{\mathrm{ExtractSign}}(x)$ without branches as follows:
+perform signed right shift of $x$ by $(w-1)$ bits.
+This leads to filling all $w$ bits with the sign bit of $x;$
+all zeros result in $0,$ while all ones result in $-1.$
+We can then use this value as a mask.
+
+For $x \in [-p; p-1],$ we define
+$\underline{\mathrm{AdjustSigned}}(x)$ as follows:
+$$
+\underline{\mathrm{AdjustSigned}}(x) =
+\begin{cases}
+x, & \text{if } x \ge 0; \\
+x+p, & \text{if } x < 0.
+\end{cases}
+$$
+
+It can be computed without branches as follows:
+$$\underline{\mathrm{AdjustSigned}}(x) = x + (p \mathbin{\&} \underline{\mathrm{ExtractSign}}(x)),$$
+where $\mathbin{\&}$ denotes bitwise ``AND''.
+Note that $p \mathbin{\&} 0 = 0,$ $p \mathbin{\&} -1 = p.$
+
+The modular addition of $a,b \in [0; p-1]$ can then be expressed as
+$$\underline{\mathrm{AdjustSigned}}(a+b-p).$$
+The normal, ``unsigned'' adjustment of $a \in [0; 2p-1]$ is expressed as
+$$\underline{\mathrm{AdjustSigned}}(a-p).$$
+The modular subtraction of $a,b \in [0; p-1]$ is expressed as
+$$\underline{\mathrm{AdjustSigned}}(a-b).$$
+
+\subsubsection{Conditional moves}
+
+The x86 and x86-64 platforms provide ``conditional moves'' --- instructions that
+store a value in a register conditionally, depending on a flag.
+They do not alter control flow and are not subject to branch prediction.
+Various versions of ARM have similar capabilities, although differently named.
+
+Since version 3.8 (latest 11.0), the Clang compiler has built-in called
+\verb~__builtin_unpredictable(x)~. According to the documentation,
+it ``is used to indicate that a branch condition is unpredictable by hardware mechanisms such as
+branch prediction logic''. We used it to implement modular addition, subtraction, and adjustment,
+which resulted in 8---13\% speedup of the overall multiplication procedure.
+
+Curiously, \verb~__builtin_unpredictable~ does not seem to produce any effect at all:
+replacing \verb~__builtin_unpredictable(x)~ with \verb~(x)~ does not affect the generated code
+at all. This means that Clang is smart enough to replace the branch with a conditional move
+even without our hint. But this does not mean, however, that we should just write the code with
+the branch and rely on the compiler to figure it out; compiling the code with branches with GNU
+GCC 10.2.0 resulted in code 40---60\% slower than the one compiled with Clang, whilst with
+the ``bit wizardry'' approach, GNU GCC outputs code with performance comparable to Clang.
+This is because GNU GCC does not rewrite the branch into a conditional move, assuming, instead,
+that it can be predicted by the hardware.
+
+The above should mean, if anything, that compiler optimizations and heuristics are very flacky,
+and should never be relied upon.
 
@@
 $\sigma_i < (B-1)M \le L.$ But then
-$$\sigma_i + c_i < 2L < \mu^2 / 2.$$
+\begin{equation} \label{eq:two_word_bound}
+\sigma_i + c_i < 2L < \mu^2 / 2.
+\end{equation}
+We see that it fits into two machine words.
 
@@
 $\frac{L}{(B-1)M}=B-1$ is going to be much larger than 1, meaning that the exact upper bound on
-$c_i + \sigma_i$ is going to be closer to $\mu^2 / 4$ than to $\mu^2 / 2;$ but this is still a
-tight bound in the sense that the maximum possible $c_i + \sigma_i$ would overflow one machine word.
+$\sigma_i + c_i$ is going to be closer to $\mu^2 / 4$ than to $\mu^2 / 2.$
+But this is still a tight bound in the sense that the maximum possible $\sigma_i + c_i$ would
+not fit into a single machine word.
 This bound also has the advantage that it works for any $B > 1.$
 
+\subsubsection{Division by constant}
+
+The iterative process~\eqref{eq:sigmai_and_zi} requires calculating both quotient and remainder
+of integer division.
+By~\eqref{eq:two_word_bound}, the dividend generally occupies two machine words.
+The divisor is $B,$ which, for our choice of bases (see section~\ref{section:bases}),
+always fits into one machine word.
+
+It is possible the built-in two-word types: \verb~uint64_t~ on 32-bit systems and
+\verb~unsigned __int128~ on 64-bit systems. But the latest versions of both Clang and GNU GCC do no
+optimize the division of two-word types by a constant, as they do for one-word types. Thus,
+this built-in division is slow.
+
+We use the approach described in~\cite{div_by_const} of rewriting a division by a constant
+into a sequence of cheaper operations, generally a single double-width multiplication,
+some bit shifts, and, for some divisors, addition and subtraction.
+For each possible value of $B$ (see, again, section~\ref{section:bases}: we support 3 different
+bases on 32-bit systems and 4 different bases on 64-bit systems), we generate the code for answer
+recovery that performs division by $B.$
+
+\subsubsection{Chinese remainder theorem}
+
+The explicit solution of finding the remainder $m = ( x \mymod (p_1 \cdot p_2) ),$
+where
+\begin{subequations}
+\begin{align}
+x \mymod p_1 &= a_1; \\
+x \mymod p_2 &= a_2,
+\end{align}
+\end{subequations}
+is given by
+\begin{equation}
+m = a_1 + p_1 \cdot \Big( \big( r \cdot (a_2 - a_1) \big) \, \mymod \, p_2 \Big),
+\end{equation}
+where
+$$r \equiv p_1^{-1} \, (\mymod \, p_2).$$
+
+If we sort the primes so that $p_1 < p_2,$
+then $( (a_2 - a_1) \mymod p_2 )$
+is just a subtraction modulo $p_2.$
+We define $\psi = (a_2 - a_1) \mymod p_2.$
+
+Note that we can pre-calculate the value of $r.$ We can also pre-calculate
+the Montgomery representation (see section~\ref{section:montgomery_intro})
+of $r$ for the second modulo,
+$$
+\widetilde{r} = (R \cdot r) \mymod p_2.
+$$
+Then, the value of $( (r \cdot \psi) \mymod p_2 )$ can be calculated as
+$$\varphi = \redc(\widetilde{r}, \psi),$$
+where the $\redc$ is done for the modulo $p_2.$
+
+The value of $m = a_1 + p_1 \cdot \varphi$ is then computed as usual.
+
 \subsection{Shenanigans with Montgomery reduction}
 
-Remember the procedure $\mathrm{MontgomeryReduce}(a)$ that we defined in section~\ref{section:montgomery_intro}.
+Remember the procedure $\mathrm{MontgomeryReduce}(a)$ that we defined in
+section~\ref{section:montgomery_intro}.
 

### 2020-11-22 paper/paper.tex
Update

@@
 \usepackage[ruled]{algorithm2e}
+\usepackage{graphicx,xcolor}
+\usepackage{gnuplottex}
+\usepackage{float}
 
@@
     Specifically, we focus on the number-theoretic transform (NTT) family of algorithms.
-    Using only portable techniques, we achieve a 3x---5x speedup over the \textbf{mpdecimal} library.
+    Using only portable techniques, we achieve a 3x---5x speedup over the \textbf{mpdecimal}
+    library.
     In this paper we describe our implementation and discuss further possible optimizations.
-    Our contributions are the following:
-    (1) the ``linearity trick'' for using Montgomery reduction with NTT without separate steps of conversion into/out of Montgomery form;
-    (2) a simple cache-efficient algorithm for in-place $2n \times n$ or $n \times 2n$ matrix transposition, the need for which arises in the ``six-step algorithm'' variation of the matrix Fourier algorithm, and which does not seem to be widely known.
-    Notable findings include:
-    (1) use of two prime moduli instead of three makes sense even considering the worst case of increasing the size of the input, and makes for simpler answer recovery;
-    (2) the ``six-step algorithm'', supposedly fast for large transforms due to cache-friendliness, is of little use on modern systems for transforms of sizes up to $\approx 2^{27}.$
+    We also present a simple cache-efficient algorithm for in-place $2n \times n$ or
+    $n \times 2n$ matrix transposition, the need for which arises in the ``six-step algorithm''
+    variation of the matrix Fourier algorithm, and which does not seem to be widely known.
+    Another finding is that use of two prime moduli instead of three makes sense even considering
+    the worst case of increasing the size of the input, and makes for simpler answer recovery.
 \end{abstract}
@@
 In order to get a provably correct answer, you have to put less bits of information into the
-floating-point values than otherwise possible~\cite{gmp_paper, mattcomp_precision}.
+floating-point values than otherwise possible
+(tables of maximum number of decimal/binary digits are given in
+\cite[p. 560--561]{mattcomp} and \cite{gmp_paper};
+for explicit formulas, see \cite{precision_paper}).
 Because of that, in order to achieve performance
@@
 Remember also that the ``high compositeness'' of $N,$ which must divide $\gcd(p_1 - 1, \ldots, p_n - 1),$ is required
-for the Cooley-Tukey algorithm to work on a length-$N$ transform. What a coincidence!
+for the Cooley-Tukey algorithm to work on a length-$N$ transform. Note the coincidence!
 
@@
 DIF and the initial permutation step in DIT.
-But note this does not mean just one of them can be omitted but not the other!
+Note this does not mean just one of them can be omitted but not the other.
 
@@
 scalar $N^{-1}$ as a separate final stage; we can ``mix in'' the factor of $R$ into this stage
-essentially for free!
+essentially for free.
 
@@
 algorithms, meaning that they interpret sequences of length $N = N_1 N_2$ as $N_1 \times N_2$
-matrices.
+matrices~\cite[p. 438--439]{mattcomp}.
 
@@
 
-DFT of length $N = RC$ can be computed as follows~\cite{mattcomp_mfa}:
+DFT of length $N = RC$ can be computed as follows:
 \begin{enumerate}
@@
 it would require $Cn$ additional memory.
+
+\subsection{Algorithm for bit-reversal permutation}
+
+See~\cite{bit_rev_perm} for overview of algorithms for bit-reversal permutation.
+We use the ``XOR'' approach presented therein, which is simple and has competitive performance
+for small-to-medium sequence sizes.
 
@@
 
-It is possible the built-in two-word types: \verb~uint64_t~ on 32-bit systems and
-\verb~unsigned __int128~ on 64-bit systems. But the latest versions of both Clang and GNU GCC do no
-optimize the division of two-word types by a constant, as they do for one-word types. Thus,
-this built-in division is slow.
+It is possible to use the built-in two-word types, \verb~uint64_t~ on 32-bit systems and
+\verb~unsigned __int128~ on 64-bit systems, for this division;
+but neither the latest Clang nor GNU GCC do optimize the division of two-word types by a
+constant, as they both do for one-word types. Thus, this built-in division is slow.
 
@@
 $$
-\widetilde{r} = (R \cdot r) \mymod p_2.
-$$
+\widetilde{r} = (R \cdot r) \mymod p_2,
+$$
+where $R = 2^w \in \mathbb{N}.$
 Then, the value of $( (r \cdot \psi) \mymod p_2 )$ can be calculated as
@@
 
+The idea for this optimization is due to~\cite{shenanigans}.
+
 Remember the procedure $\mathrm{MontgomeryReduce}(a)$ that we defined in
 section~\ref{section:montgomery_intro}.
-
-\subsection{Cache aliasing problems}
+It requires $$0 \le a < \mu p,$$ where $p$ is our prime modulo,
+$\mu = 2^w,$ $w$ is the length of machine word in bits.
+In order to compute $\redc(x,y),$ we call $\mathrm{MontgomeryReduce}(xy).$
+Normally, we have $0 \le x,y < p.$
+By~\eqref{eq:pi_bound}, we also have $p < \mu/2.$
+It means that $$a = xy < p \mu / 2.$$
+
+This means that we can call $\mathrm{MontgomeryReduce}(xy)$ if either $x$ or $y$ (but not both)
+is in ``unadjusted'' form. Being ``unadjusted'' means being in $[0; 2p-1],$
+thus possibly having different representation than the ``canonical'' one for this value
+(in $[0; p-1]$).
+
+We can obtain such an ``unadjusted'' value by omitting an adjustment:
+\begin{itemize}
+    \item After an addition: the unadjusted addition of $x,y \in [0; p-1]$ is just $x + y.$
+    \item After a subtraction: the unadjusted subtraction of $x,y \in [0; p-1]$ is $x - y + p.$
+    \item After $\redc$: observe that $\mathrm{MontgomeryReduce}$ itself
+          performs adjustment as the last step.
+\end{itemize}
+
+Note that we use this optimization much more sparingly than we theoretically could, because
+allowing non-local propagation of ``unadjustment'' would be a nightmare from the standpoint of
+reliability and debuggability: we would need to keep track of what can, and what can not, be
+``unadjusted'' at each step.
+
+\section{Prospects}
+
+We see the following possible ways to further improve the performance of decimal multiplication.
+
+\begin{enumerate}
+  \item \textbf{Improve cache utilization}.
+    \begin{enumerate}
+      \item Use breadth-first ordering, perhaps using explicit recursion.
+            This can be beneficial even considering the function call overhead,
+            as it leads to better cache utilization~\cite{fftw}.
+      \item Explore the ``Belgian approach''; see~\cite{belgian}.
+      \item Explore using higher-radix transforms (raidx-4, radix-8, etc).
+            Note that radix-4 is beneficial for complex FFT because,
+            for a standard representation by a pair of floating-point values,
+            multiplication by $i$ or $-i$ can be done significantly faster than normal
+            multiplication; but finite fields of integers modulo $p$ with standard representation
+            lack this property.
+            Nevertheless, higher-radix transforms might be of service because they have better cache
+            locality~\cite{fftw}.
+    \end{enumerate}
+  \item \textbf{Additional shenanigans}.
+    \begin{enumerate}
+      \item It is suggested in \cite{shenanigans, mont_sp} to use primes $p$ such that $p < \mu/4.$
+          Then, the Montgomery multiplication can be implemented in such a way that if the inputs
+          are in $[0; 2p-1],$ then the outputs are also in $[0; 2p-1],$ and no final conditional
+          subtraction is performed.
+      \item It is stated in~\cite{shenanigans_presentation} that we could benefit from
+          higher-radix transforms by further eliminating reductions from the butterflies.
+      \item Explore signed Montgomery reduction, as defined in~\cite{signed_montgomery}.
+    \end{enumerate}
+  \item \textbf{Smooth the stairs}, possibly utilizing ideas from~\cite{smooth_the_stairs}.
+        Currently, we can only perform transforms of lengths $2^k$ and $3 \cdot 2^k.$
+        It would be nice to be able to perform transforms of lengths that are in-between.
+\end{enumerate}
+
+\section{Benchmark}
+
+\begin{figure}[H]
+    \centering
+    \begin{gnuplot}[terminal=epslatex, terminaloptions=color]
+        set style line 1 lt 1 lw 2 lc rgb '#4682b4' pt -1
+        set style line 2 lt 1 lw 2 lc rgb '#ee0000' pt -1
+
+        set xtics axis
+        set ytics axis
+
+        set grid xtics lc rgb '#555555' lw 1 lt 0
+        set grid ytics lc rgb '#555555' lw 1 lt 0
+
+        set ytics 0.5
+        set xtics 1
+
+        set xlabel "$log_2(n)$"
+        set ylabel "ratio"
+
+        set autoscale xfix
+
+        plot \
+            'benchdata.txt' using 1:2 with lines ls 1 ti "CMOVs on", \
+            'benchdata.txt' using 1:3 with lines ls 2 ti "CMOVs off"
+    \end{gnuplot}
+\end{figure}
 

### 2020-11-22 paper/paper.tex
Update

@@
+% ``Speeding up decimal multiplication'' (c) by shdown
+%
+% ``Speeding up decimal multiplication'' is licensed under a
+% Creative Commons Attribution 4.0 International License.
+%
+% You should have received a copy of the license along with this
+% work. If not, see <http://creativecommons.org/licenses/by/4.0/>.
+
 \documentclass[a4paper,12pt]{article}
@@
 
-Unfortunately, this method is very slow; in our benchmark, native division on x86-64
-was 720\% slower than the Montgomery reduction. The code can be found in the
-\verb!bench-modmul! subdirectory of our repository.
+Unfortunately, this method is very slow; see section~\ref{section:modmul_bench} for the
+results of our benchmark against Montgomery reduction and Solinas reduction.
 
@@
 We found that on a modern x86-64 system, reduction with two rounds of $r$ is slower by
-$\approx$ 15\% compared to Montgomery reduction;
-the code can be found in the \verb!bench-modmul! subdirectory of our repository.
+$\approx$10\% compared to Montgomery reduction;
+see section~\ref{section:modmul_bench}.
 
@@
 single-word $p=2^{N_1}-2^{N_2}+1$ useless, if Montgomery reduction can be used instead.
+
+\subsubsection{Benchmark}
+\label{section:modmul_bench}
+
+We benchmarked naïve division using x86-64
+\verb~divq~
+instruction, Montgomery reduction,
+and Solinas reduction.
+
+Given an argument $n,$ our benchmark performs $n$ iterations of loop with
+$10$ unrolled modular multiplications, performing in total $10 n$ modular multiplications.
+
+We compiled it using Clang 11.0.0, with option \verb!-O3!. The machine is
+Xiaomi RedmiBook 14" 2019 JYU4203CN laptop with
+Intel\textsuperscript{\textregistered} Core\texttrademark{} i3-8145U CPU @ 2.10GHz;
+CPU scaling governors for all CPUs were set to ``performance''.
+
+\begin{center}
+\begin{tabular}{ |c|c|c| }
+    \hline
+        Approach & $n$ & Time, s \\
+    \hline\hline
+        Naïve & $10^8$ & 23.04 \\
+    \hline
+        Montgomery & $10^8$ & 2.86 \\
+    \hline
+        Solinas & $10^8$ & 3.19 \\
+    \hline
+\end{tabular}
+\end{center}
+
+The code can be found in the \verb!bench-modmul! subdirectory of our repository
+(see section~\ref{section:avail}).
 
@@
 \subsubsection{Bit wizardry}
+\label{section:bitwiz}
 
@@
 \subsubsection{Conditional moves}
+\label{section:cmovs}
 
@@
             as it leads to better cache utilization~\cite{fftw}.
-      \item Explore the ``Belgian approach''; see~\cite{belgian}.
+      \item Explore what~\cite{gmp_paper} call the ``Belgian approach''; see~\cite{belgian}.
       \item Explore using higher-radix transforms (raidx-4, radix-8, etc).
@@
 
+We benchmarked our implementation (see section~\ref{section:avail} for the link to the repository)
+against \textbf{mpdecimal}.
+
+Since both rely on Cooley-Tukey,
+the graphs of time as a function of input size $n$ (in decimal digits)
+would be staircase-like.
+Define ``threshold input size'' for some fixed implementation as any value of $n$ where the
+discontinuity occurs --- a new ``stair'' pops up.
+
+We calculated the threshold input sizes of both our implementation and of the \textbf{mpdecimal}
+library, within reasonable limits ($2176 \le n \le 3\cdot10^7$).
+We merged all threshold values into a single list, sorted it, then formed a new list from the
+averages of each two adjacent values in the sorted list.
+We then appended the maximum threshold value plus one to the resulting list.
+It is easy to see that this way nothing gets lost --- assuming that the staircase, aside from
+discontinuities, is horizontal, we measure all possible cases.
+
+For each value of $n,$ we performed $\lfloor \frac{8 \cdot 10^7}{n} \rfloor$
+multiplications of numbers of $n$ decimal digits each.
+
+We compiled everything using Clang 11.0.0, with option \verb!-O3!. The machine is
+Xiaomi RedmiBook 14" 2019 JYU4203CN laptop with
+Intel\textsuperscript{\textregistered} Core\texttrademark{} i3-8145U CPU @ 2.10GHz;
+CPU scaling governors for all CPUs were set to ``performance''.
+
+We used \textbf{mpdecimal} version 2.5.0-4 from Debian Bullseye repositories
+for the ``amd64'' architecture.
+For further information on \textbf{mpdecimal}, refer to
+its homepage \url{http://www.bytereef.org/mpdecimal/index.html}.
+
 \begin{figure}[H]
@@
     \begin{gnuplot}[terminal=epslatex, terminaloptions=color]
+
         set style line 1 lt 1 lw 2 lc rgb '#4682b4' pt -1
@@
         set xlabel "$log_2(n)$"
-        set ylabel "ratio"
+        set ylabel "Ratio"
 
@@
             'benchdata.txt' using 1:3 with lines ls 2 ti "CMOVs off"
+
     \end{gnuplot}
@@
 
+``Ratio'' is time of \textbf{mpdecimal} divided by time of our implementation.
+The ``CMOVs on'' variant uses Clang-specific
+\verb~__builtin_unpredictable~ built-in for modular addition, subtraction, and adjustment
+(see section~\ref{section:cmovs});
+the ``CMOVs off'' variant relies on ``bit wizardry'' instead (see section~\ref{section:bitwiz}).
+
+\section{Availability}
+\label{section:avail}
+
+The code of our implementation, benchmark scripts, and \LaTeX source of this paper,
+are available at \url{https://github.com/shdown/decimal-multiplication-paper}.
+The code is licensed under the MIT license.
+The source of this paper is licensed under the Creative Commons BY 4.0 license.
+
 \bibliography{paper}{}

### 2020-11-22 bench-modmul/bench.c

@@ -1,3 +1,6 @@
+// (c) 2020 shdown
+// This code is licensed under MIT license (see LICENSE.MIT for details)
+
 #include <stdint.h>
 #include <stdio.h>
 #include <stdlib.h>
@@ -103,18 +106,17 @@ int main(int argc, char **argv)
     }
     ULIMB x = parse_int_or_die(argv[1]);
     int n = parse_int_or_die(argv[2]);
-    ULIMB y = x >= 23 ? (x - 23) : (x + 23);
     for (int i = 0; i < n; ++i) {
-        x = mul(x, y);
-        y = mul(y, x);
-        x = mul(x, y);
-        y = mul(y, x);
-        x = mul(x, y);
-        y = mul(y, x);
-        x = mul(x, y);
-        y = mul(y, x);
-        x = mul(x, y);
-        y = mul(y, x);
+        x = mul(x, x);
+        x = mul(x, x);
+        x = mul(x, x);
+        x = mul(x, x);
+        x = mul(x, x);
+        x = mul(x, x);
+        x = mul(x, x);
+        x = mul(x, x);
+        x = mul(x, x);
+        x = mul(x, x);
     }
-    printf("%u, %u\n", (unsigned) x, (unsigned) y);
+    printf("%u\n", (unsigned) x);
 }

### 2020-11-22 bench.c

@@ -1,3 +1,6 @@
+// (c) 2020 shdown
+// This code is licensed under MIT license (see LICENSE.MIT for details)
+
 #include "fft.h"
 #include <stdint.h>
 #include <stdlib.h>

### 2020-11-22 bench_gmp.c

@@ -1,52 +1,0 @@
-#include <gmp.h>
-#include <stdio.h>
-#include <stdlib.h>
-#include <time.h>
-
-static int x_parse_int(const char *s)
-{
-    int r;
-    if (sscanf(s, "%d", &r) != 1) {
-        fprintf(stderr, "Cannot parse '%s' as integer.\n", s);
-        abort();
-    }
-    return r;
-}
-
-int main(int argc, char **argv)
-{
-    if (argc != 4) {
-        fprintf(stderr, "USAGE: bench_gmp BASE NDIGITS NREPEAT\n");
-        return 2;
-    }
-    int base = x_parse_int(argv[1]);
-    int ndigits = x_parse_int(argv[2]);
-    int nrepeat = x_parse_int(argv[3]);
-
-    mpz_t x;
-    mpz_init(x);
-    mpz_ui_pow_ui(x, base, ndigits);
-    mpz_sub_ui(x, x, 1);
-
-    mpz_t y;
-    mpz_init_set(y, x);
-    mpz_sub_ui(y, y, 1);
-
-    mpz_t z;
-    mpz_init(z);
-
-    fprintf(stderr, "base=%d, ndigits=%d, nrepeat=%d\n", base, ndigits, nrepeat);
-    clock_t t = clock();
-
-    for (int i = 0; i < nrepeat; ++i) {
-        mpz_mul(z, x, y);
-    }
-
-    t = clock() - t;
-    printf("%.4f\n", ((double) t) / CLOCKS_PER_SEC);
-
-    mpz_clear(x);
-    mpz_clear(y);
-    mpz_clear(z);
-    return 0;
-}

### 2020-11-22 bench_mpdec.c

@@ -1,3 +1,6 @@
+// (c) 2020 shdown
+// This code is licensed under MIT license (see LICENSE.MIT for details)
+
 #include <mpdecimal.h>
 #include <stdio.h>
 #include <stdlib.h>

### 2020-11-22 fft.c

@@ -1,3 +1,6 @@
+// (c) 2020 shdown
+// This code is licensed under MIT license (see LICENSE.MIT for details)
+
 #include "fft.h"
 #include "fft_utils.h"
 #include "fft_compdep.h"

### 2020-11-22 fft.h

@@ -1,3 +1,6 @@
+// (c) 2020 shdown
+// This code is licensed under MIT license (see LICENSE.MIT for details)
+
 #pragma once
 
 #include <stdint.h>

### 2020-11-22 fft_compdep.h

@@ -1,3 +1,6 @@
+// (c) 2020 shdown
+// This code is licensed under MIT license (see LICENSE.MIT for details)
+
 #pragma once
 
 #include <stddef.h>

### 2020-11-22 fft_utils.c

@@ -1,3 +1,6 @@
+// (c) 2020 shdown
+// This code is licensed under MIT license (see LICENSE.MIT for details)
+
 #include "fft_utils.h"
 #include "fft_compdep.h"
 

### 2020-11-22 fft_utils.h

@@ -1,3 +1,6 @@
+// (c) 2020 shdown
+// This code is licensed under MIT license (see LICENSE.MIT for details)
+
 #pragma once
 
 #include "fft.h"

### 2020-11-22 gen/fastdiv.py

@@ -1,3 +1,6 @@
+# (c) 2020 shdown
+# This code is licensed under MIT license (see LICENSE.MIT for details)
+
 from utils import ceil_log2, decompose_pow2
 
 

### 2020-11-22 gen/gen.py

@@ -1,4 +1,8 @@
 #!/usr/bin/env python3
+
+# (c) 2020 shdown
+# This code is licensed under MIT license (see LICENSE.MIT for details)
+
 import argparse
 
 from isprime import is_prime

### 2020-11-22 gen/isprime.py

@@ -1,3 +1,6 @@
+# (c) 2020 shdown
+# This code is licensed under MIT license (see LICENSE.MIT for details)
+
 from utils import decompose_pow2, mod_pow
 
 

### 2020-11-22 gen/utils.py

@@ -1,3 +1,6 @@
+# (c) 2020 shdown
+# This code is licensed under MIT license (see LICENSE.MIT for details)
+
 def egcd(a, b):
     s = 0
     old_s = 1

### 2020-11-22 main.c

@@ -1,3 +1,6 @@
+// (c) 2020 shdown
+// This code is licensed under MIT license (see LICENSE.MIT for details)
+
 #include "fft.h"
 #include <stdint.h>
 #include <inttypes.h>

### 2020-11-22 main.c
Update

@@ -8,10 +8,6 @@
 #include <stdio.h>
 #include <string.h>
 
-#ifndef METHOD
-#   define METHOD 1
-#endif
-
 static void die_oom(void)
 {
     fprintf(stderr, "Out of memory.\n");
@@ -46,15 +42,21 @@ static size_t div_ceil_zu(size_t x, size_t y)
     return x / y + !!(x % y);
 }
 
-static size_t x_calc_fft_size(size_t m)
+static size_t x_calc_fft_size(unsigned method, size_t m)
 {
-#if METHOD == 1 || METHOD == 6
-    size_t n = 1;
-#elif METHOD == 4
-    size_t n = 3;
-#else
-#   error "Unsupported METHOD."
-#endif
+    size_t n;
+    switch (method) {
+    case 1:
+    case 6:
+        n = 1;
+        break;
+    case 4:
+        n = 3;
+        break;
+    default:
+        fprintf(stderr, "Unknown method, expected either of: 1, 4, 6.\n");
+        abort();
+    }
 
     while (n < m)
         n = x_mul_zu(n, 2);
@@ -155,8 +157,22 @@ static int max_base_log(size_t ndigits)
     }
 }
 
-int main()
+int main(int argc, char **argv)
 {
+    if (argc != 2) {
+        fprintf(stderr, "USAGE: main METHOD\n");
+        fprintf(stderr, "  where METHOD must be either of: 1 4 6.\n");
+        fprintf(stderr, "METHOD of 1 means use the 'straight' FFT (function 'fft()').\n");
+        fprintf(stderr, "METHOD of 4 means use the four-step FFT (function 'fft_fourstep()').\n");
+        fprintf(stderr, "METHOD of 6 means use the six-step FFT (function 'fft_sixstep()').\n");
+        return 2;
+    }
+    unsigned method;
+    if (sscanf(argv[1], "%u", &method) != 1) {
+        fprintf(stderr, "Cannot parse '%s' as unsigned integer (METHOD).\n", argv[1]);
+        return 2;
+    }
+
     char *a = read_number_str();
     size_t na = strlen(a);
 
@@ -172,7 +188,7 @@ int main()
     size_t nlimbs_a = div_ceil_zu(na, base_log);
     size_t nlimbs_b = div_ceil_zu(nb, base_log);
[…]

### 2020-11-22 paper/paper.tex
Update

@@
 
-The code of our implementation, benchmark scripts, and \LaTeX source of this paper,
+The code of our implementation, benchmark scripts, and \LaTeX{}
+source of this paper,
 are available at \url{https://github.com/shdown/decimal-multiplication-paper}.

### 2020-11-22 paper/paper.tex
Update

@@
 By $\mathbb{N}$ we mean $\{0, 1, 2, \ldots\}.$
+
+By $x \mymod y,$ where $x \in \mathbb{Z},$ $y \in \mathbb{N} \setminus \{0\},$
+we mean $z \in \mathbb{N} \cap [0; y-1]$ such that $z \equiv x \, (\mymod \, y).$
+
+By $x \mydiv y,$ where $x \in \mathbb{N},$ $y \in \mathbb{N} \setminus \{0\},$
+we mean $\lfloor x / y \rfloor.$
 
@@
 where $C_1$ and $C_2$ denote loop-invariant expressions that can be calculated once before the loop.
-But on x86 and x86-64, there is difference:
-these platforms can encode SIB (scale-index-base) directly in the instruction that dereferences the
+But on x86 and x86-64 platforms, there \textit{is} a difference:
+they can encode SIB (scale-index-base) directly in the instruction that dereferences the
 pointer, but only for constant scales of 1, 2, 4 and 8 bytes.
@@
 
-The x86 and x86-64 platforms provide ``conditional moves'' --- instructions that
+The x86-64 platform provides ``conditional moves'' --- instructions that
 store a value in a register conditionally, depending on a flag.

### 2020-11-23 paper/paper.tex
Update

@@
             as it leads to better cache utilization~\cite{fftw}.
-      \item Explore what~\cite{gmp_paper} call the ``Belgian approach''; see~\cite{belgian}.
+      \item Explore what~\cite{gmp_paper} calls the ``Belgian approach''; see~\cite{belgian}.
       \item Explore using higher-radix transforms (raidx-4, radix-8, etc).
@@
     \begin{enumerate}
-      \item It is suggested in \cite{shenanigans, mont_sp} to use primes $p$ such that $p < \mu/4.$
+      \item It is suggested in \cite{shenanigans, mont_sp} that we use primes $p$ such that $p < \mu/4.$
           Then, the Montgomery multiplication can be implemented in such a way that if the inputs

### 2020-11-23 paper/paper.tex
Update

@@
 \textbf{mpdecimal} would use it.
-After all, the case of six-step FFT with its
+The case of six-step FFT with its
 $2^k \times 2^{k+1}$ or $2^{k+1} \times 2^k$ matrices must be an important enough
@@
 
-Now, the case of inverse transform.
-We now need to multiply each matrix element indexed $\langle i,j \rangle$
+Consider now the case of inverse transform.
+We need to multiply each matrix element indexed $\langle i,j \rangle$
 by $(\xi_N)^{-ij},$ then perform length-3 IDFT on each column; and also multiply everything by a

### 2020-11-25 paper/paper.tex
Update

@@
 
-\begin{procedure}[H]
+\begin{procedure}[!hbt]
 \DontPrintSemicolon
@@
 
-\begin{procedure}[H]
+\begin{procedure}[!hbt]
 \caption{FindPrimesForN($n$, $p_{\max}, \ell$)}
@@
 
-\begin{procedure}[H]
+\begin{procedure}[!hbt]
 \caption{FindPrimes($w$, $n_{\min}$, $\ell$)}
@@
 
-\begin{procedure}[H]
+\begin{procedure}[!hbt]
 \caption{BitRevPermute($A$)}
@@
 
-\begin{procedure}[H]
+\begin{procedure}[!hbt]
 \caption{DecimationInTime($A$, $\omega$)}
@@
 
-\begin{procedure}[H]
+\begin{procedure}[!hbt]
 \caption{DecimationInFrequency($A$, $\omega$)}
@@
 matrix.
-For completeness, we provide the code below.
-
-\begin{procedure}[H]
+For completeness, we provide the code of $\mathrm{TransposeSquareSubMatrix}(p, n, n').$
+For a cache-oblivious version of this algorithm, see~\cite{cache_square_mat_trans}.
+
+\begin{procedure}[!hbt]
 \caption{TransposeSquareSubMatrix($p$, $n$, $n'$)}
@@
 \end{procedure}
-
-For a cache-oblivious version of this algorithm, see~\cite{cache_square_mat_trans}.
 
@@
 
-We provide the pseudocode for case of $C=2$ below.
-
+We provide the pseudocode for case of $C=2.$
 Note it uses the routine $\mathrm{TransposeSquareSubMatrix}$.
@@
 
-\begin{procedure}[H]
+\begin{procedure}[!hbtp]
 \caption{PermuteRho($p$, $\zeta$, $n$)}
@@
 
-\begin{procedure}[H]
+\begin{procedure}[!hbtp]
 \caption{UnPermuteRho($p$, $\zeta$, $n$)}
@@
 
-\begin{procedure}[H]
+\begin{procedure}[!hbtp]
 \caption{TransposeMatrixNx2N($p$, $\zeta$, $n$)}
@@
 
-\begin{procedure}[H]
+\begin{procedure}[!hbtp]
 \caption{TransposeMatrix2NxN($p$, $\zeta$, $n$)}

### 2020-11-25 paper/paper.tex
Update

@@
 
+    \item IDFT* means IDFT without multiplication by $N^{-1}.$
+
 \end{itemize}
@@
 
-IDFT of length $N = RC$ can be computed as follows, not counting final multiplication by $N^{-1}$:
+IDFT* of length $N = RC$ can be computed as follows:
 \begin{enumerate}
     \item $[C \times R],$ transpose the matrix.
-    \item $[R \times C],$ perform (length $C$) IDFT on each row with $\xi = (\xi_N)^R.$
+    \item $[R \times C],$ perform (length $C$) IDFT* on each row with $\xi = (\xi_N)^R.$
     \item $[R \times C],$ multiply each element indexed $\langle i,j \rangle$ by $(\xi_N)^{-ij}.$
-    \item $[R \times C],$ perform (length $R$) IDFT on each column with $\xi = (\xi_N)^C.$
+    \item $[R \times C],$ perform (length $R$) IDFT* on each column with $\xi = (\xi_N)^C.$
 \end{enumerate}
@@
         and column index $j,$ where $0 \le j < N_2,$ is said to be indexed $\langle i, j \rangle.$
+
+    \item IDFT* means IDFT without multiplication by $N^{-1}.$
 
@@
 
-IDFT of length $N = RC$ can be computed as follows, not counting final multiplication by $N^{-1}$:
+IDFT* of length $N = RC$ can be computed as follows:
 \begin{enumerate}
     \item $[C \times R],$ transpose the matrix.
-    \item $[R \times C],$ perform (length $C$) IDFT on each row with $\xi = (\xi_N)^R.$
+    \item $[R \times C],$ perform (length $C$) IDFT* on each row with $\xi = (\xi_N)^R.$
     \item $[R \times C],$ multiply each element indexed $\langle i,j \rangle$ by $(\xi_N)^{-ij}.$
     \item $[R \times C],$ transpose the matrix.
-    \item $[C \times R],$ perform (length $R$) IDFT on each row with $\xi = (\xi_N)^C.$
+    \item $[C \times R],$ perform (length $R$) IDFT* on each row with $\xi = (\xi_N)^C.$
     \item $[C \times R],$ transpose the matrix.
@@
 the same property can be employed to get rid of the
-final transposition step in DFT and the initial transposition step in IDFT simultaneously.
+final transposition step in DFT and the initial transposition step in IDFT* simultaneously.
 
@@
 \item multiplication by $(\xi_N)^{-ij},$
-\item length-3 IDFT on each column,
+\item length-3 IDFT* on each column,
 \end{itemize}
@@
 We need to multiply each matrix element indexed $\langle i,j \rangle$
-by $(\xi_N)^{-ij},$ then perform length-3 IDFT on each column; and also multiply everything by a
+by $(\xi_N)^{-ij},$ then perform length-3 IDFT* on each column; and also multiply everything by a
 factor $\Phi.$

### 2020-11-27 paper/paper.tex
Update

@@
 \date{%
-    \today
+    November 25, 2020
 }
@@
 It is a common theme in practical analysis of algorithms that ``fancy'' algorithms are slow when $n$ is small.
-For example, virtually all partical implementations of the ``fancy'' sorting algorithms that are divide-and-conquer in nature --- quick sort and merge sort --- do in fact fall back to ``dumb'' quadratic sorts if the size of the input if less than some threshold.
+For example, virtually all partical implementations of the ``fancy'' sorting algorithms that are divide-and-conquer in nature --- quick sort and merge sort --- do in fact fall back to ``dumb'' quadratic sorts if the size of the input is less than some threshold.
 In some cases, it is not even a matter of ``fancy'' versus ``dumb'' dichotomy, but rather there is a ``hierarchy of fanciness'':
@@
 it is based on the use of prime moduli of form $2^{64} - 2^{n} + 1.$
-Generally, such primes are knows as Solinas primes, or generalized Mersenne
+Generally, such primes are known as Solinas primes, or generalized Mersenne
 primes~\cite{Solinas},
@@
 We use the Miller-Rabin primality test with the first 12 prime numbers as bases; it has been proven
-in \cite{primebases} that, for values less than $2^{64},$ this is enough to guarantee correctness.
+in~\cite{primebases} that, for values less than $2^{64},$ this is enough to guarantee correctness.
 
@@
 if we use DIF for the direct transforms in definition of four-step DFT, and DIT for inverse
-transforms in the definition of four-step IDFT.
+transforms in the definition of four-step IDFT*.
 
@@
 
-We provide the pseudocode for case of $C=2.$
+We provide the pseudocode for the case of $C=2.$
 Note it uses the routine $\mathrm{TransposeSquareSubMatrix}$.
@@
 Note that we use this optimization much more sparingly than we theoretically could, because
-allowing non-local propagation of ``unadjustment'' would be a nightmare from the standpoint of
+allowing non-local propagation of ``unadjustment'' would be a nightmare from the standpoints of
 reliability and debuggability: we would need to keep track of what can, and what can not, be
@@
 would be staircase-like.
-Define ``threshold input size'' for some fixed implementation as any value of $n$ where the
+Define ``threshold input size'' for some fixed implementation as any value of $n$ where a
 discontinuity occurs --- a new ``stair'' pops up.

### 2020-12-09 paper/paper.tex
Update

@@
 This means $c_1 p \equiv 1 \, (\mymod \, \mu)$ and $-c_1 p \equiv -1 \, (\mymod \, \mu),$ so
-$$p' = (-c_1 p) \mymod \mu.$$
+$$p' = (-c_1) \mymod \mu.$$
 
